## Logical Vectors: Create with Repetitions and Summary Functions {#logical-patterns}

Whenever you need to create repeated vectors that are of logical type, you can conveniently achieve such tasks through methods you learned in \@ref(patterned_numeric_vectors_creation) and \@ref(char-concatenate).

### Create logical vectors with repetitions

Let's start by recalling how we create numeric and character vectors with repetitions. The function we use is the `rep()` function, inside of which the first argument refers to elements to be repeated, and the second argument indicates the number of repetitions. 

```{r}
num1 <- rep(1, 3)
cha1 <- rep("sheep", 2)
```

Remember that, the first argument can be 1) a number, 2) a character, 3) colon operator `:` (for numeric vectors), 4) a numeric vector with several values, and 5) a character vector with several values; the second argument can be 1) a number, 2) colon operator `:`, and 3) a numeric vector with several values. When only the first argument is a vector with several values, it is **repeated as a whole**, not element-wisely. When both arguments are vectors with several values, R will do an **element repeat** operation.

```{r}
num2 <- rep(c(1, 0), 3)
num3 <- rep(1:3, 3:1)
num4 <- rep(c(1, 0), c(2, 2))
char2 <- rep(c("apple", "chocolate"), 2)
char3 <- rep(c("apple", "chocolate"), 1:2)
char4 <- rep(c("apple", "chocolate"), c(3, 3))
```

If you feel confident creating numeric and character vectors with repetitions, the ways of creating logical vectors with repetitions are basically at your fingertips:

```{r eval=FALSE}
logic <- rep(c(TRUE, FALSE, NA), c(3,2,1))
logic
```


### Statistical Functions on Logical Vectors {#functions-logi}

Before we end our study of logical vectors with logical operators in next section, we will take a look at how some previously introduced summary functions can be used on logical vectors.

***Group A: summary statistics***

What if we apply `summary()` on logical vectors?

```{r}
logic <- rep(c(TRUE, FALSE, TRUE, NA), 4:1)
summary(logic)
```

Similar to character vectors, you can get the vector type, which is `logical` here. You also get a frequency table for the times `FALSE`, `TRUE`, and missing values `NA` appear in the vector.

***Group B: arithmetic operations***

Different from character vectors, logical vectors can be used in ordinary arithmetic, meaning you can apply almost all the functions summarized in Table \@ref(tab:summaryFuns) to logical vectors. In such cases, the **coercion rule** introduced in Section \@ref(coercion-rule) will convert every `TRUE` to 1 and `FALSE` to 0. Let's take a look at an example,

```{r}
a <- c(TRUE, TRUE, FALSE, FALSE, TRUE)
sum(a)
mean(a)
```

Clearly, to proceed calculations, R considers `a` having three 1s and two 0s, given there are three `TRUE` and two `FALSE`. Therefore, sum(a)` equals 3 and `mean(a)` equals 0.6. You are welcome to try other functions on a logical vector. 

However, as much as `TRUE` and `FALSE` will be coerced into specific numeric values, missing values `NA` won't be coerced to an assigned value. Therefore, ordinary arithmetic involving `NA` will return `NA` as well, meaning the result is **not available**.

```{r}
b <- c(TRUE, FALSE, TRUE, NA)
sum(b)
mean(b)
```

```{r}
d <- c(1:5, NA)
sum(d)
e <- c(1:5, NaN)
sum(e)
f <- c(1:5, NA, NaN)
is.na(d)
is.nan(d)
```



### Exercises

1. Write the R code to create logical vectors with the following results, then show the unique elements and their corresponding frequencies in each vector:

a).`FALSE TRUE NA FALSE TRUE NA`
b). `FALSE FALSE TRUE TRUE TRUE NA`


2. Write two different R codes using `rep()` function to create a logical vector with values `TRUE FALSE TRUE FALSE`. The two codes should have at least one argument different from each other.









