## Missing Values (`NA`) {#missing-values}

So far, we have been working with the ideal case that the vector is fully observed without any missing values. However, in most applications, you may encounter the situation, where some values like `NA` or a large value `999`, which represent **missingness** in the data set for certain variables, and such values can be resulted from one of the below situations:

a) the observer fails to record the data, meaning the value is not available at the given position of a vector or a data frame, or 
b) the data is impossible or undefined, meaning the value is resulted from an error or logical contradiction, such as dividing by 0.

In two scenarios, R uses a) `NA`, indicating they are **not available**, and b) `NaN`, indicating they are **not a number** to present those values, respectively. Let's see the following example. 

```{r, results = TRUE}
a <- 1:10
a[11]
```

Here, you have defined `a` as a length-10 numeric vector with 10 values in it. If you try to access the 11-th element of `a`, it is not available, hence you will see `NA` as the result. 


Sometimes, the values of some elements in a vector are missing, then you can use `NA` for these elements. Here is an example containing `NA`s as some values. If you want to get values of the 2nd and 4th elements in `b`, you will get `NA NA` as the result.

```{r, results = TRUE}
b <- c(1, NA, 2, NA, 3)
b
b[c(2,4)]
```

Let's see another example.

```{r}
b <- 0/0
```

If you look at the environment panel, you will see that `b` has a value of `NaN`, because arithmetically you cannot divide a number by 0.

```{r}
c <- c(NA, NA)
class(b)
typeof(b)
class(c)
typeof(c)
```

By running the above codes, you can notice that `NA` is stored as a logical type, whereas `NaN` is stored as a numeric type.

Now you have had a basic understanding of `NA` and `NaN`. In the following parts of this section, we want to introduce several properties of them. Let's start with the fact that `NA` is **contagious**. 

### `NA` and `NaN` are contagious

***a. `NA` ***

`NA` implies that the underlying value is not available, in other words, there is uncertainty with the value. As a result, for most operations associated with `NA`, the results will also be `NA`, showing that `NA` is **contagious**. 

```{r, results=TRUE}
y <- NA
y + 3
y == 3
```

As you can see here, `y` is `NA`, indicating the value of `y` is not available. When you try to do operations like `y + 3` or `y == 3`, the answers are clearly not available as well, hence both taking the value `NA`. 

How about we create another `NA` object and compare it with `y`? 
```{r, results=TRUE}
z <- NA 
y == z 
NA == NA
```

It is again `NA`, which may be confusing at a first thought. However, keep in mind that since both `y` and `z` are not available, there is no way to tell whether they are the same. Hence `y == z` is also `NA`.

::: {.infobox .caution data-latex="{caution}"}
1. Think about what is the value of 1^NA^ in R. Try to run it in R. Does it agree with your thoughts? 

2. Think about what is the value of 0^NA^ in R. Try to run it in R. Does it agree with your thoughts? 

- Answer 1. For any $x$, we have $1^x = \exp(\log (1^x)) = \exp(x \log 1) = \exp(x \cdot 0) = 1$. Since there is no uncertainty regarding the expression, the value of 1^NA^ is 1 
- Answer 2. We have $0^0 = \lim_{x\to 0}x^x =\exp[\lim_{x\to 0} x\log(x)]=\exp[0] = 1$, and $0^1 = 0$. Since `NA` represents uncertainty values, it can be 0 or 1 or other numbers. So 0^NA^ is not deterministic because it can take different values according to the exponent. Hence, the value of 0^NA^ is also `NA`.
::: 


Now, let's talk about what impact the `NA` values make when we apply statistical functions. Let's create a vector containing `NA` values, and apply some functions on it.
```{r, results=TRUE}
x <- c(1, NA, 3, 4, NA, 2)
x
sum(x) 
mean(x)
sd(x)
```

As you can see, for many statistical functions on vectors, as long as there exists at least one `NA` values in vectors, the results are often impossible to determine, hence resulting an `NA` value as well. 

As a result, you may want to ignore the `NA` values during the function evaluation. Fortunately, most statistical functions on vectors provide an optional argument `na.rm`, which takes a logical value, indicating whether to remove `NA` before applying the functions. Let's see the following examples.
```{r, results=TRUE}
sum(x, na.rm = TRUE) 
mean(x, na.rm = TRUE)
sd(x, na.rm = TRUE)
```

It is easy to verify that the results are what we expect to get if the `NA` values are removed. Feel free to try the following codes which apply the same functions on the subvector with the non-missing values.
```{r, eval = FALSE}
x_no_na <- c(1, 3, 4, 2)
sum(x_no_na) 
mean(x_no_na)
sd(x_no_na)
```

Interesting, the `summary()` function will deal with the `NA` values automatically by removing them before computing the five percentiles and the mean. In addition, the `summary()` function provides a column which shows the number of `NA`s in `x`.
```{r, results=TRUE}
summary(x) 
```

***b. `NaN` ***

```{r}
b * 3
b + 3
```

Similar to `NA`, `NaN` is also contagious, because numeric calculations on undefined values are also undefined.

```{r}
d <- c(1, 2, 3, 0/0)
d
sum(d) 
mean(d)
sd(d)
sum(d, na.rm = TRUE) 
mean(d, na.rm = TRUE)
sd(d, na.rm = TRUE)
```

Moreover, if a vector consists undefined values, applying statistical functions on it will also result in `NaN`, and you can still use `na.rm` to remove `NaN` before applying the functions.

### Work with `NA` values

Since `NA` is far more common in real-life applications, we'll focus this section on `NA` values. When there are `NA` values in our vector, there is nothing to be afraid of as there are many useful tools we can use. 

Let's use `x <- c(1, NA, 3, 4, NA, 2)` throughout this part. 

***a. Find indices with missing values***

Firstly, we will introduce how to find indices with missing values. To find the indices, you may be tempted to use the comparison operator introduced in Section \@ref(create-logical-vector). Let's try to compare `x` with `NA` as the following code.
```{r}
x == NA
```

You get a vector of all values equaling `NA`! This is actually not surprising for the following reason. Given that the `NA` you are comparing can take any unknown value, any comparison with it will result in an `NA` value due to the lack of information.

Instead of using `==` for finding missing values, the correct way is to use the `is.na()` function, which returns a logical vector representing whether the value of each element is missing or not. Then, you can use the `which()` function which can return the locations of all `TRUE` values to find the indices for the `NA` values. Here, the `sum()` function on the logical vector `x_na` returns the number of `NA` values in the vector, following the coercion rule described in Section \@ref(functions-logi). 
```{r, results=TRUE}
x_na <- is.na(x)        #logical vector     
x_na
which(x_na)             #numeric vector
sum(x_na)               #the number of NAs in x 
sum(!x_na)              #the number of non-NAs in x
```

If you only want to detect whether there is any `NA` values in the vector, you can use the `anyNA()` function.
```{r, results=TRUE}
anyNA(c(NA, 1))
anyNA(1:3)
```


***b. Remove missing values***

Sometimes, you may want to simply remove the missing values. To do that, you can use a logical vector to do vector subsetting as introduced in Section \@ref(vector-subsetting). The specific logical vector you want to use is the opposite (`!`) of the logical vector that represents missing values. Then you will get a subvector of `x` which keeps all values except `NA`. You can also use the `na.omit()` function to achieve the same result.
```{r, results=TRUE}
x2 <- x[!x_na]
x2
na.omit(x)
```

***c. Impute missing values***

In many applications, naively removing the missing values before doing the analysis may lead to incorrect inference. Usually, it is useful to make the data complete by **imputing** the missing values. For example, you can use mean imputing or median imputing, which replaces the missing values with the mean or median of the non-missing values. 

```{r, results=TRUE}
x_impute <- x
mean(x, na.rm = TRUE)
x_impute[x_na] <- mean(x, na.rm = TRUE)
x_impute
x
```

::: {.infobox .caution data-latex="{caution}"}
If you want to compare the values of an object before and after some operations, you can create a new object with the same value as the original object (here, we create `x_impute` which has the same value as `x`), then make operations on `x_impute` without changing the value of `x`. 
::: 
In `x_impute`, values of the 2nd and 5th elements are replaced by 2.5, the average of the non-missing values of `x`. 

***d. Replace non-standard missing values with `NA` in the vector***

Sometimes, the data we collected may not use `NA` to indicate that a value is missing. For example, in the following vector `x3`, the value `999` represents the corresponding element that is missing.

```{r}
x3 <- c(4, 999, 1, 999, 3, 999, 999)
```

It is highly recommended to convert these values into `NA` before carrying out any analysis. 
```{r, results=TRUE}
x3[x3 == 999] <- NA
x3
```

Let's see another example where both `999` and `-999` represent the value is missing. You can convert all `999` and `-999` into `NA` by using operations introduced in Section \@ref(set-operations-vectors).
```{r, results=TRUE}
x4 <- c(4, 999, 1, -999, 3, -999, 999)
##999 and -999 are the values indicating missingness
x4[x4 %in% c(999, -999)] <- NA
x4
```



### Exercises
1. For the vector `x <- rep(c(1, 2, NA), 3:5)`, 
a. verify each value of `summary(x)` by using other functions.
b. find the indices with missing values;
c. create a vector x_no_na containing the non-missing values in `x`;
d. replace those missing values by the median of the non-missing values in `x`.

2. For the vector `y <- rep(c("N", 2, "A"), 5:3)`, the values of both `"N"` and `"A"` indicate missingness. Convert non-standard missing values to `NA`, then find the indices of `y` that correspond to missing values. 




