## Data Frame {#dataframe}

So far, we have learned atomic vectors (Chapter \@ref(r-objects)), matrices (Section \@ref(matrix)), and arrays (Section \@ref(array)). Regardless of their dimensionality, the three different object types share an important features: they all contain elements of the same type (i.e., numeric, character, or logical). In real applications, it is common to have mixed variable types. To accommodate this, let's introduce a new two-dimensional object type: **data frame**. 

### Introduction to Data Frames

To create a data frame, you can use the `data.frame()` function to collect several vectors of the same length. Let's see an example of some health conditions of a sheep and a pig over the years 2019, 2020 and 2021. 
```{r, results=TRUE}
animal <- rep(c("sheep", "pig"), c(3,3))
year <- rep(2019:2021, 2)
weight <- c(110, 120, 140, NA, 300, 800)
height <- c(2.2, 2.4, 2.7, 2, 2.1, 2.3)
condition <- c("excellent", "good", NA, "excellent", "good", "average")
condition <- factor(condition, ordered = TRUE, levels = c("average", "good", "excellent"))
healthy <- c(rep(TRUE, 5), FALSE)
my_data_frame <- data.frame(animal, year, weight, height, condition, healthy)
my_data_frame
```

Looking at the data frame `my_data_frame`, it has 6 columns, each of which represents one variable. The variables are of different types. The `animal` is factor, `year` is integer, both `weight` and `height` are doubles, `condition` is ordered factor, and `healthy` is logical. 

After creating the data frame, it is useful to examine its class using the `class()` function and structure using the `str()` function. 

```{r, results=TRUE}
class(my_data_frame)
str(my_data_frame)
```

The `str()` tells us the data frame has 6 observations and 6 variables, along with the type and the first few values of each variable. From the output, you may be puzzled by the `$` symbol before each variable name. In fact, you can easily extract a certain column corresponding to a variable with the `$` following by its name. 

```{r, results=TRUE}
my_data_frame$animal
my_data_frame$weight
```


::: {.infobox .caution data-latex="{caution}"}
This kind of data representation is impossible using matrices since the coercion rule will apply, converting everything into characters. Let's combine everything into a matrix and check its value. 
```{r, results=TRUE}
my_mat <- cbind(animal, year, weight, height, condition, healthy)
my_mat
```
::: 

In the process of creating data frames, you can also name each column.
```{r, eval=FALSE}
my_data_frame2 <- data.frame(ani = animal, y = year, w = weight, h = height, con = condition, hea = healthy)
```


In Section \@ref(numeric-vector-statistics), we introduced the very useful function `summary()`, which returns important summary statistics for a vector. Using `summary()` on a data frame, you get the summary statistics for each variable. 

```{r, results=TRUE}
summary(my_data_frame)
```

From the results, you can see that depending on the variable type, you get different forms of summary. 

In real world, it is very common to encounter missing values, and you may want to jettison those missing values or observations with them. In the object `my_data_frame`, there are two missing values represented by `NA`. To remove the observations (rows) with `NA` values, you can use the `na.omit()` on the data frame.
```{r, results=TRUE}
my_df_nona <- na.omit(my_data_frame)
my_df_nona
```

You can see that the 3rd and 4th row are removed since they both have a missing observation. 

An alternative approach to remove all rows with missing observations is to first use the `complete.cases()` function to get a logical vector of whether a row has missing elements, and then use data frame subsetting. 
```{r}
complete_ind <- complete.cases(my_data_frame)
my_data_frame[complete_ind, ]
```



### Adding Observations or Variables in Data Frames

Sometimes, you may want to add additional entries to the 1st dimension (i.e., rows/observations) or the 2nd dimension (i.e., columns/variables) to an existing data frame. 

To add additional observations, you need to put the additional observations or variables into a new data frame, and use the `rbind()` function. 

```{r, results = TRUE}
new_obs <- data.frame(animal = "pig", year = 2018, weight = 200, height = 1.9, condition = "excellent", healthy = TRUE)
rbind(my_data_frame, new_obs)
```

To add an additional variable to the existing data frame, you just need to use the `$` operator to name the new variable and use the `<-` operator to assign values to the new variable.


```{r}
my_data_frame$age <- my_data_frame$year - 2015
my_data_frame
```

### Subsetting and Modifying Data Frames {#subset_df}

As a two-dimensional object type, subsetting and modifying data frames is very similar to the operations on matrices. 

***a. using indices to do data frame subsetting and modifying***

The first method for data frame subsetting is to specify the desired row indices and column indices, separated by `,`. For example, we can extract the (1, 3) and (2, 4) element of `x` using the following codes. 
```{r, results=TRUE}
my_data_frame[1, 3]
my_data_frame[2, 4]
```

To modify a specific element of `x`, we can assign the desired value to it.
```{r}
my_data_frame[1, 3] <- 115
```

You can check the correponding value of the data frame is modified.

```{r, eval=FALSE}
my_data_frame
```


Similar to a matrix subsetting, if you omit the indices of one dimension, R will keep everything along that dimension. You can also use negative indices to keep everything except the provides indices. Let's see some examples. 
```{r, results=TRUE}
my_data_frame[2, ]
my_data_frame[, 3]
my_data_frame[3]       #a single index corresponds to the columns
my_data_frame[c(1,3), -c(2, 3, 4)]
```

Similarly, we can modify multiple values at the same time.
```{r}
my_data_frame[1:3, 3] <- c(120, 123, 141)
my_data_frame
```

Note that you generally want to supply a vector with the same length as the number of elements to be modified, or the recycling rule will apply.

***b. using column names to do data frame subsetting and modifying***

Since data frames usually have column names, you can do subsetting using multiple column names. 
```{r, results = TRUE}
my_data_frame[1:2, c("animal", "weight")]
```

Let's try to modify this sub data frame.

```{r}
my_data_frame[1:2, c("animal", "weight")] <-
  data.frame(animal = c("monkey", "bear"),
             weight = c(110, 300))
my_data_frame
```


***c. using logical vectors to do data frame subsetting and modifying***

Using logical vectors to do data frame subsetting can come in handy. Suppose we want to find the condition of the pig in year 2021. 
```{r, results=TRUE}
is_2021 <- my_data_frame$year == 2021
is_pig <- my_data_frame$animal == "pig"
my_data_frame[is_2021 & is_pig, "condition"]
```

Now, let's say we want to extract all the observations with an excellent condition. 
```{r, results=TRUE}
my_data_frame[my_data_frame$condition == "excellent", ]
my_data_frame[which(my_data_frame$condition == "excellent"), ]    #remove the NA row
```

Let's try to modify the condition of the pig in year 2021 to "excellent".
```{r}
my_data_frame[is_2021 & is_pig, "condition"] <- "excellent"
```

### Exercises
Consider the following data frame,
```{r, results=TRUE}
animal <- rep(c("sheep", "pig"), c(3,3))
weight <- c(110, NA, 140, NA, 300, 800)
condition <- c("excellent", "good", NA, "excellent", "good", "average")
healthy <- c(rep(TRUE, 5), FALSE)
my_data_frame <- data.frame(animal, weight, condition, healthy)
my_data_frame
```

1. Generate a data frame with rows of `my_data_frame` containing complete observations.
2. In `my_data_frame`, fill in the missing values in `weight` by the median of its non-missing values. 
3. Add the following observation to `my_data_frame`: `animal = "pig"`, `weight = 900`, `condition = average`, and `healthy = FALSE`. 
4. Extract the sub-data-frame of `my_data_frame` that contains the columns of `animal` and `healthy` and the rows that has `weight` less than 400 and condition is `"good"` or `"excellent"`.
