## NULL, NaN, and Inf {#null}

Having learned the special missing value representation `NA`  in Section \@ref(missing-values), we will introduce three additional values to represent unexpected results, namely the `NULL`, `NaN`, and `Inf`. During the process, we will talk about their relationships to `NA` as well. All `NULL`, `NaN`, and `Inf` are reserved words in R.

Before introducing the three types, let's review the structure, internal storage type, and length of `NA`. It is worth recalling that `NA` is a length-1 logical constant that indicates a missing value. Keep in mind that `NA` is different from `"NA"` (a character vector).

```{r, results=TRUE}
str(NA)
typeof(NA)
length(NA)
```

### NULL

First, let's take a look at `str()`, `typeof()` and `length()` of `NULL`. 
```{r, results=TRUE}
str(NULL)
typeof(NULL)
length(NULL)
```


From the output, you'll see that `NULL` is a length-0 object often returned by expressions and functions whose value is undefined. 

***a. Undefined field of a list***

The first scenario of `NULL` is when you try to access an element of a list that is undefined. 
```{r, results = TRUE}
my_list <- list(num = 1:3, char = c("a", "b"))
my_list$logi
```
Here, the result is `NULL` since logi is not a defined field in `my_list`. 

***b. Remove an element from a list***

You can remove an element from a list by assign it the `NULL` value.
```{r, results=TRUE}
length(my_list)
my_list$num <- NULL
length(my_list)
my_list
```

As you can see from the output, the element `num` is removed from `my_list`, leading to the length of `my_list` reduced by 1.

***c. Initialize a list of certain length***

The `NULL` value is useful to serve as the default initial value when you want to create a list of certain length using the `vector()` function. 

```{r, results=TRUE}
my_list <- vector(mode = "list", length = 3)
my_list
```

It is worth mentioning that the `vector()` function is also useful to initialize a vector of given mode and length. 
```{r, results = TRUE}
vector("numeric", length = 2)      ##default is 0
vector("logical", length = 2)      ##default is FALSE
vector("integer", length = 2)      ##default is 0
vector("character", length = 2)    ##default is empty string
```

To check if an element is `NULL`, you can't use the logical comparison `== NULL`. Instead, you need to use the `is.null()` function.

```{r, results = TRUE}
a <- NULL
a == NULL
is.null(a)
```

It is worth explaining the result of `a == NULL` is `logical(0)`, representing a logical vector of length 0. The underlying reason is that `NULL` contains no value and is of length 0. As the `==` comparison returns a logical type object, hencing leading to a logical vector of length 0. 


***d. `NULL` values when creating a vector***

If you create a vector with `NULL` values, all `NULL` values will be removed if there exists at least one defined values. If all of them are `NULL` values, only one of them will be kept, but that vector's length will be 0. Note that there is fundamentally different from `NA` values. `NA` means the value is there, but the exact value is not available to us. 

```{r, results=TRUE}
c(NULL, NULL, 1, NULL)
c(NULL, NULL)
c(NA, NA)
```

One final point on `NULL` is that you can technically perform calculations on `NULL`. However, from the example below, you'll realize operations on `NULL` will only change the class, leaving the length unchanged.

```{r}
d <- NULL
d <- d + 5
str(d) # the class is changed to numeric, but the length is still 0
```

### NaN {#NaN}

`NaN`, represents **Not a Number**, usually appears when you divide 0 by 0, indicating the result is not well-defined.
```{r, results = TRUE}
0/0
```

Again, it is worth to look at `str()`, `typeof()` and `length()` of `NaN`.

```{r, results = TRUE}
str(NaN)
typeof(NaN)
length(NaN)
```

As you can see from the results, `NaN` is a numeric vector of length 1, with the value `NaN`. 

To check if a value is `NaN`, you cannot use the `== NaN` similar to checking missing values, instead you need to use the function `is.nan()`.
```{r, results=TRUE}
a <- NaN
a == NaN       ##resulting an NA value
is.nan(a)      ##the correct way to check if the value is NaN
is.nan(c(NA, 1, NaN))
```


### Inf {#Inf}

The last special we want to introduce in this section is `Inf`, representing the value is positive **infinity** ($\infty$), corresponding to a proper mathematical limit. Similarly, we also have negative infinity: `-Inf`. 
```{r, results = TRUE}
1/0
-2/0
Inf > 3
Inf < -1
Inf + Inf
-Inf + 1e10
1/0 - 1/0        #it equals 0/0, hencing NaN
```

Again, it is worth to look at `str()`, `typeof()` and `length()` of `Inf`.

```{r, results = TRUE}
str(Inf)
typeof(Inf)
length(Inf)
```

As you can see from the results, similar to `NaN`, `Inf` is a numeric vector of length 1, but with the value `Inf`. 

To check whether a value is finite or infinite, you can use the `is.finite()` and `is.infinite()` function. 
```{r, results=TRUE}
is.finite(1/0)
is.infinite(-3/0)
```

### A comparison of the four special values in R

We would like to summarize the different behaviors of the four special values in R in the following table. 

Summary   |`NA`      | `NULL`     | `NaN`      | `Inf`
----------|----------|------------|------------|--------------
`class()` |`"logical"` | `"NULL"`     |`"numeric"`   |`"numeric"`
`length()`|   1      |   0        |   1        |  1
check     |`is.na()` | `is.null()`| `is.nan()` | `is.finite()`

### Exercises

Suppose `x <- c(NA, NULL, Inf, NaN)`, answer the following questions. (Try to solve without using R first and then confirm with R code)

1. What's the length of `x`?
2. What's the class and the storage type of `x`?
3. What's the value of `x + 1`? Explain the reason for each element in the result.
4. What's the value of `x == x`? Explain the reason for each element in the result.
