## ggplot Aesthetics (I): Constant-Valued Aesthetics {#aes-constant-value}

Having learned how to generate a basic scatterplot using `geom_point()` in Section \@ref(scatterplot), we are now ready to introduce one of the most important ingredients in a `geom`, namely, the **aesthetics**. Aesthetics include distinct features that you can change that affect the appearances of a plot, including color, shape, and size.

There are in general three kinds of aesthetics:

1. **Constant-Valued Aesthetics**: this type of aesthetics change the corresponding feature of a plot **globally** (for example, `color = "red"` will make everything red). This type of aesthetic specification appears as **an argument of the geoms** (for example `geom_point()`). This will be the focus of our current section. 

2. **Local Aesthetic Mapping**: this type of aesthetics (for example, `color`) will take possibly different values (different color values) for different data points depending on the **mapped variables, only for the current geom**. This type of aesthetic specification appears as an argument of the `aes()` function of the corresponding geom. We will cover this type of aesthetic in the next Section (Section \@ref(aes-local-map)).

3. **Global Aesthetic Mapping**: this type of aesthetics will use possibly different values for different data points depending on the **mapped variable, at the global level**, which will be **passed** to all individual geoms. This type of aesthetic specification appears as an argument of the `aes()` function, which is an argument of the `ggplot()` function. We will cover this type of aesthetic in Section \@ref(multi-geom), where multiple geoms are used. 

::: {.infobox .caution data-latex="{caution}"}
Note that although we will introduce aesthetics via the example of scatterplot, they are used for all kinds of plots which will be covered at a later time. 
::: 


Let's first review the code we used to generate the scatterplot between `sugar` and `cholestrol` in the `gm2004` dataset.

```{r, results = TRUE}
library(ggplot2)
library(r02pro)
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, y = cholesterol))
```

Now, let's see how to set **Constant-Valued Aesthetic** in `geom_point()`. 

***a. Color***

To change the color of all points, you can add a `color` argument in the `geom_point()` function. Note that it is placed **outside** of the `aes()` function, which is different from aesthetic mappings, to be covered in the next section.

```{r, results = TRUE}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, y = cholesterol),
             color = "blue")
```

Clearly, all points are changed to blue. 

***b. Size***

Similarly, you can set the `size` aesthetic in the `geom_point()` function to change the size of the all points.

```{r, results = TRUE}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, y = cholesterol),
             size = 3)
```

Note that it is very common to set multiple constant valued aesthetic in a geom function.

```{r, results = TRUE}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, y = cholesterol),
             color = "blue",
             size = 3)
```

You may notice that the points are now bigger than before. Looking at the plot, some points are overlapping with each other, which is sometimes called **overplotting**. To solve this issue, you can change the transparency level of the points by setting the `alpha` aesthetic. 

***c. Transparency***

```{r, results = TRUE}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, y = cholesterol),
             color = "blue",
             size = 3, 
             alpha = 0.5)
```

By setting `alpha = 0.5`, the points become more visible and the overplotting problem is largely alleviated. 

***d. Shape***

Lastly, we can also change the shape of the points from the default `1` (circle) to other shapes by the `shape` argument in `geom_point()`. 

```{r, results = TRUE}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, y = cholesterol),
             color = "blue",
             size = 3, 
             alpha = 0.5,
             shape = 2)
```


Now, we have all points blue, size of 3, of triangle shape, and of transparency level 0.5. Recall the collection of all shapes is available in Figure \@ref(fig:all-shapes).

You are welcome to try different combinations of global aesthetics.


### Exercises


1.	Using the `sahp` dataset, create a scatterplot with **ggplot2** to visualize the relationship between `lot_area` (on the x-axis) and `sale_price` (on the y-axis) with all points of color purple and size 2.   

2. Using the `gm2004` dataset, create a scatterplot with **ggplot2** to visualize the relationship between `alcohol` (on the x-axis) and `liver_cancer` (on the y-axis) with all points of color pink, size 3, transparency 0.3, and shape of diamond. 




