## Smoothline Fits {#smoothline}

Now, you know how to create scatterplots and line plots with many possible customizations via specifying different aesthetics. In addition to scatterplots, a very useful type of plots that can capture the trend of pairwise relationship is the **smoothline fits**.

### Creating Smoothline Fits using `geom_smooth()`

To create a smoothline fit, you can use the `geom_smooth()` function in  the **ggplot2** package. Let's say you want to find the trend between cholesterol and sugar in the `gm2004` dataset. 

```{r}
library(ggplot2)
library(r02pro)
ggplot(data = gm2004) + 
  geom_smooth(mapping = aes(x = sugar, 
                           y = cholesterol))
```

Perhaps it is helpful to review the code for generating a scatterplot between cholesterol and sugar. 

```{r, eval = FALSE}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol))
```

We can see that the **only** difference between these two codes is the use of different geoms. In fact, the mechanism of `geom_smooth()` is that it fits a smooth line according to the points of the given variable pair. By default, it uses the **loess** method (locally estimated scatterplot smoothing), which is a popular nonparametric regression technique. In addition to the smoothline, it also generates a shaded area, representing the **confidence interval** around the fitted smoothline. 




### Constant-Valued Aesthetics in Smoothline Fits

Just like in scatterplots (Section \@ref(aes-constant-value)), you can also set Constant-Valued Aesthetics for smoothline fits. 

***a. Confidence Interval (Hide or Display)***

As mentioned before, the default smoothline fit display a confidence interval around the smooth curve. To hide this shaded area, you can add the argument `se = FALSE` as a Constant-Valued aesthetic. 

```{r}
ggplot(data = gm2004) + 
  geom_smooth(mapping = aes(x = sugar, 
                           y = cholesterol),
              se = FALSE)
```

***b. Confidence Interval Level***

The default confidence interval is at 95% level. To control the level, you can set the aesthetic `level` to a desired value.

```{r}
ggplot(data = gm2004) + 
  geom_smooth(mapping = aes(x = sugar, 
                           y = cholesterol),
              level = 0.9)
```


***c. Smoothing Method***

In addition to the default loess method for smoothline fit, `geom_smooth()` also provides other smoothing methods. We can set the `method` aesthetic to change the smoothing method.  For example, `method = "lm"` represents a linear line fit.

```{r}
ggplot(data = gm2004) + 
  geom_smooth(mapping = aes(x = sugar, 
                           y = cholesterol),
              method = "lm")
```

***d. Color***

Similar to Scatterplot, you can also set `color` to change the color of the line. Note that the `color` aesthetic is available for almost **all** geoms. 

```{r}
ggplot(data = gm2004) + 
  geom_smooth(mapping = aes(x = sugar, 
                           y = cholesterol),
              color = "violet")
```

***e. Size***

Similar to Scatterplot, you can also set the `size` aesthetic. While the `size` controls the size of the points in a scatterplot, the `size` aesthetic controls the **width** of the smoothline. 

```{r}
ggplot(data = gm2004) + 
  geom_smooth(mapping = aes(x = sugar, 
                           y = cholesterol),
              size = 3)
```

***f. Line Type***

Another useful aesthetic that was not applicable in `geom_point()` is `linetype`, which controls the linetypes for each smoothline. The collection of different linetypes is available in Figure \@ref(fig:all-line-types).

```{r}
ggplot(data = gm2004) + 
  geom_smooth(mapping = aes(x = sugar, 
                           y = cholesterol),
              linetype = 2)
```


Note that it is common to use multiple constant-valued aesthetics. 

```{r}
ggplot(data = gm2004) + 
  geom_smooth(mapping = aes(x = sugar, 
                           y = cholesterol),
              color = "purple",
              size = 2,
              method = "lm",
              level = 0.99,
              linetype = 2)
```

This generates a smoothline fit with purple color, line width 2, linear smoothing, 0.99 confidence level, and dashed line.

### Local Aesthetics Mapping in Smoothline Fits

As in scatterplots (Section \@ref(aes-local-map)), you can also map variables to aesthetics in smoothline fits. 

***a. Group***

When we map a variable to the `group` aesthetic, `geom_smooth` will first divide all the data points into different *groups* according to the variable value, and then fit a separate smoothline for each group. 

```{r}
ggplot(data = gm2004) + 
  geom_smooth(mapping = aes(x = sugar, 
                            y = cholesterol,
                            group = gender))
```

You can see that two smoothlines are generated. However, it is not clear from the plot which group each smoothline corresponds to. To better differentiate the two groups, you can map the variable to other aesthetics like `color`. 

***b. Color***

As in `geom_point()`, we can map the variable to the color aesthetic. 

```{r}
ggplot(data = gm2004) + 
  geom_smooth(mapping = aes(x = sugar, 
                            y = cholesterol,
                            color = gender))
```

This is a more informative plot than the one using `group` aesthetic as you can see the two smoothlines have different colors according to the value of `gender`. We can observe that when the value of sugar is low, it appears that female has a larger cholesterol value than male and the difference becomes smaller as the sugar value increases.


***c. Line Type***

We can also use different line types to differentiate a variable by the `linetype` aesthetic.

```{r}
ggplot(data = gm2004) + 
  geom_smooth(mapping = aes(x = sugar, 
                            y = cholesterol,
                            linetype = gender))
```

The plot shows a dashed line for the smoothline corresponding to `gender = "male"`, and a solid line for the smoothline corresponding to `gender == "female"`.


***d. Size***

You can also map `gender` to the `size` aesthetic, which controls the width of each smoothline fit.

```{r}
ggplot(data = gm2004) + 
  geom_smooth(mapping = aes(x = sugar, 
                            y = cholesterol,
                            size = gender))
```

Note that similar to `geom_point()`, you can see a warning message: "Using size for a discrete variable is not advised." 

It is worth to mention that `shape` is not a valid aesthetic for `geom_smooth` as it doesn't make sense to talk about the shape of a line. 

```{r, error = TRUE}
ggplot(data = gm2004) + 
  geom_smooth(mapping = aes(x = sugar, 
                            y = cholesterol,
                            shape = gender))
```

When you try to map a variable to the `shape` aesthetic, `geom_smooth()` will show a warning message "Warning: Ignoring unknown aesthetics: shape", and use the `group` aesthetic instead. 




Naturally, you can mix different local aesthetic mappings as well as different constant-valued aesthetics on the same plot. 
```{r}
ggplot(data = gm2004) + 
  geom_smooth(mapping = aes(x = sugar, 
                            y = cholesterol,
                            color = gender,
                            linetype = gender),
              size = 2,
              se = FALSE)

```



### Exercises

Using the `sahp` dataset with the **ggplot2** package, answer the following questions.


1.	Create a smoothline fit to visualize the relationship between `lot_area` (on the x-axis) and `sale_price` (on the y-axis).
2.  Create several smoothlines with different colors corresponding to the value of `kit_qual` to visualize the relationship between `lot_area` (on the x-axis) and `sale_price` (on the y-axis).
2.  Create smoothlines without confidence interval around and with different linetypes to distinguish whether the house has more than 2 bedrooms to visualize the relationship between `lot_area` (on the x-axis) and `sale_price` (on the y-axis) .







