## ggplot Aesthetics (II): Local Aesthetics Mapping {#aes-local-map}

In Section \@ref(aes-constant-value), we introduced the first kind of aesthetics, namely **Constant-Valued Aesthetics**, where we set constant values to aesthetics that change features of the plot globally. In some applications, however, you may want to highlight different groups of the data with different values of aesthetics. In this section, we will dive deep into this topic: **Local Aesthetics Mapping**.

Believe it or not, we've already seen this kind of aesthetic mapping when we first introduced ggplot in Section \@ref(point).

```{r}
library(ggplot2)
library(r02pro)
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol)) 
```

In the `aes()` function, `x` and `y` can also be viewed as aesthetics. It is clear to see that we are not setting `x` and `y` as constants. Instead, we map the variable `sugar` to `x` and the variable `cholesterol` to `y`. The `x` and `y` coordinates control the location of each data points depending on the variables of `sugar` and `cholesterol`, which is the essence of scatterplot. Now, let's see how to do local aesthetic mappings for color, size, shape, and so on. 

### Basic Local Aesthetics Mapping

***a. Color***

Let's say you want to use different colors for `male` and `female` on the scatterplot. To do this, you can map the discrete variable `gender` to the `color` aesthetic by setting `color = gender` as an argument in the `aes()` function.

```{r}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           color = gender))
```

From this figure, we can clearly see that different genders are shown in distinct colors. Also, for the same value of `sugar`, it appears that the value of `cholestrol` could be a bit smaller for male than that for female. In addition, `ggplot()` automatically created a legend to show the correspondence between the genders and colors.

Can we map a continuous variable to the color aesthetic? The answer is positive.

```{r scatterplot-color}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           color = population))
```

Here, you can see that the color of the points changes smoothly according to the value of `population`, the details of which is shown in the legend. In particular, the darker points represent smaller populations while the lighter ones correspond to bigger populations.


***b. Size***

In addition to color, you can also map a discrete variable to the size aesthetic. 

```{r}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           size = gender),
             alpha = 0.5)
```

You can see from the plot that the sizes of the points are now different according to the `gender`. 


To alleviate the overplotting issue, we added a **contant-valued aesthetic** `alpha = 0.5` to the `geom_point()` function, making all points more transparent.  Please pay close attention to the location where we put the argument:

1. Contant-Valued Aesthetic: the aesthetic specification is put inside the `geom_point()` function.
2. Local Aesthetics Mapping: the aesthetic mapping is put inside the `aes()` function, which is itself an argument of the `geom_point()` function. 
 


There is a warning message: "Using size for a discrete variable is not advised." The reason is that different sizes may implicitly indicate a particular ordering of the groups, which are usually not clear for a discrete variable. 

Now, let's try to map a continuous variable `BMI` to the size aesthetic. 

```{r}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           size = BMI),
             alpha = 0.5)
```

Note that now the size of the points changes according to the corresponding `BMI` value, i.e. the larger points represent higher BMIs values and the smaller ones correspond to lower BMI values. It is worth noting that the legend only shows three representative `BMI` values (20, 25, an 30) and their corresponding sizes, however, we have other sizes show on the plot depending on exact BMI values. 

***c. Shape***

We can also map a discrete variable to the `shape` aesthetic. 

```{r}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           shape = continent))
```

Now, the shape of the points are different for each continent. 

Can we map a continuous variable to the shape aesthetic? Let's try it out.

```{r, error=TRUE, eval=FALSE}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           shape = BMI))
```

You will see an error message: "A continuous variable can not be mapped to shape". This is because shape is of a discrete nature in the sense that different shapes have no natural ordering and the number of possible shapes is  finite where the continuous variable could take infinitely many different values. 

***d. Multiple Aesthetic Mappings***

Just like Constant Valued Aesthetics, you can have multiple aesthetic mappings in ggplot, which can involve multiple mapped variables.

```{r}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           shape = gender,
                           color = HDI_category))
```

Here, we differentiate `gender` by different shapes and differentiate `HDI_category` by different colors. Note that `HDI_category` represent the Human Development Index for each country. We can see that countries with low HDI has smaller values of sugar and cholesterol in general. Also note that there are two legends on the plot showing the color and shape, respectively.

### Customize the Local Aesthetics Mapping

So far, we have introduced the basic local aesthetics mapping, where the default aesthetic values are used. Sometimes, you may want to use specific aesthetic values (colors, shapes) for particular values of the mapped variable. 

***a. Customize colors for discrete variables***

Let's start with the color aesthetic. To customize the colors with discrete variable, you can add a layer to the ggplot with function `scale_colour_manual` with  argument `values` containing a character vector consisting of the desired colors in order.

```{r}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           color = gender)) + 
  scale_colour_manual(values = c("orange","purple"))
```

Now, the female group has color orange and the male group has color purple. 


***b. Remove the `NA` level***

Let's try to map `continent` to `color`. 

```{r}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           color = continent))
```

The plot shows us the Africa countries in general have a lower value of `sugar` as well as a small value of `cholestrol`. On the other hand, the American and European countries have large values of `sugar` and `cholestrol`. 

You may noticed that this is a level `NA` for the continent due to missing values. To address this issue, you can use the `remove_missing()` function to remove observations with a missing value of `continent`. 

```{r}
ggplot(data = remove_missing(gm2004, vars = "continent")) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           color = continent))
```

Now, only the five continents are displayed. 

::: {.infobox .caution data-latex="{caution}"}
Note that here, we add the argument `vars = "continent"` in the `remove_missing()` function. This works by removing the observations only if the `continent` variable is missing. Without such an argument, an observation will be removed as long as there is at least one variable missing. For a dataset like `gm2004` that contain many variables, we recommend explicitly specify the variables that need to be considered for missingness. Here, the `vars` can take the value of a character vector, containing all variables that you want to consider for missingness.
::: 

***c. Customize color for continuous variables***

To customize the color for continuous variables, you can add a layer `scale_color_continuous()` which specify the lower-value color and the high-value color for the variable. 

```{r color-discrete-customize}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           color = BMI)) + 
  scale_color_continuous(low = "green", 
                         high = "red")
```

Now, we mapped the `BMI` to the color aesthetic and the lower values of `BMI` are mapped to green while the higher values are mapped to red. 

***d. Customize the shape for discrete variables***

Similarly, you can also customize the shapes for each level of the discrete variable. 

```{r}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           shape = gender)) +
  scale_shape_manual(values = c(6, 3))
```

Here, we have the female and male being mapped to shape 6 and 3, respectively. Recall that all possible shapes are shown in Figure \@ref(fig:all-shapes).


### Customize the legend order for discrete variables

Let's review the following plot where we map `continent` to the color aesthetic.

```{r}
ggplot(data = remove_missing(gm2004, vars = "continent")) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           color = continent))
```

Looking at the legend, you can see that different `continent` values are in alphabetical order as introduced in Section \@ref(character-sort-rank-order) when you learned the ordering of character vectors. Sometimes, you may want to arrange these values in a different order in the plot. To achieve this, you can add a layers `scale_color_discrete()` with the argument `breaks` that specifies the desired order on the legend.

```{r}
ggplot(data = remove_missing(gm2004, vars = "continent")) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           color = continent))  +
  scale_color_discrete(breaks = c("Oceania", "Europe", "Asia", "Americas","Africa"))
```

Note that only the legend order changed with the corresponding mapped values unchanges. 

Now, let's map `gender` to the shape aesthetic.
```{r}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           shape = gender))
```

To customize the legend order, you can add another layers called `scale_shape_discrete()` with the desired legend order.

```{r}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           shape = gender)) +
  scale_shape_discrete(breaks = c("male", "female"))
```

You may find a pattern here that to customize a ggplot where a certain aesthetic is considered, we need to use the corresponding customization function like `scale_color_discrete()` for the color aesthetic and `scale_shape_discrete()` for the shape aesthetic. 

### Create Logical Variables and Map to Aesthetics

Lastly, you can also create logical variables on the fly and map them to aesthetics, without defining them as new variables. For example, if you want to differentiate the points according to whether the value of `BMI` is larger than 25, a logical variable `BMI > 25` can be created.

```{r}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           color = BMI > 25),
             size = 2)
```

Now, let's see another example where we want differentiate countries in Africa vs. other continents. 

```{r}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           shape = continent != "Africa"),
             size = 2)
```


Clearly, you can easily create new logical variables using any logical operations on existing variables, and map them into any aesthetics just like the existing categorical variables.

### Exercises

Use **ggplot2** package to answer the following questions.

1.	Using the `sahp` dataset, create a scatterplot to visualize the relationship between `liv_area` (on the x-axis) and `sale_price` (on the y-axis).  Use different colors for different `house_style`, different shapes to distinguish `central_air`, and points are of the size 2.

2. Using the `gm2004` dataset, create a scatterplot to visualize the relationship between `alcohol` (on the x-axis) and `liver_cancer` (on the y-axis) where the points are purple and the size of the points are determined by the `BMI` values. 

3. Using the `gm2004` dataset, create a scatterplot to visualize the relationship between `sugar` (on the x-axis) and `cholesterol` (on the y-axis) where the points are red when `HDI_category` is high or very high and the points are purple when `HDI_category` is low or medium and the size depend on the value of `population`. Please also remove the `NA` level in `HDI_category`. Discuss your findings.






