## Bar Charts and Pie Charts via geom_bar() {#bar-charts}

For visualizing the relationship between two **continuous variables**, we have learned various kinds of plots, including scatterplot (Sections \@ref(scatterplot) - \@ref(aes-local-map)), line plot (Section \@ref(line-plot)), and smoothline fit (Section \@ref(smoothline)), 

In this section, we will introduce two new kinds of plots for **discrete variables**, called **bar chart** and **pie chart**. A bar chart uses rectangular bars with heights or lengths proportional to the values they represent, in order to visualize a discrete variable. A pie chart divides a circle into slices to represent numerical proportions.

### An Introduction to Bar Chart {#intro-bar-chart}

In the `sahp` dataset, you may be interested in the distribution of kitchen quality of the house (denoted in the `kit_qual` variable). To generate a bar chart, you can use the function `geom_bar()`. 

```{r}
library(ggplot2)
library(r02pro)
ggplot(data = sahp) + 
  geom_bar(mapping = aes(x = kit_qual))
```

In the bar chart, the x-axis displays different values of `kit_qual`, and the y-axis displays the number of observations with each `kit_qual` value. To verify this, let's check how many house have `kit_qual` equals "Excellent". 
```{r, results = TRUE}
sum(sahp$kit_qual == "Excellent")
```

We can see the answer `r sum(sahp$kit_qual == "Excellent")` matches the value on the bar chart. 

You may have noticed that the y-axis count is not a variable in `sahp`! This is also the reason that we **don't** need to specify the `y` argument in the `aes()` function. In this sense, bar charts are very different from many other graphs like scatterplots, which plot the raw values of datasets. 



Sometimes, we would like to find out the proportion for each value of `x`, then we can display a bar chart of proportion, rather than count. To do this, we need to add `y = after_stat(prop)` and `group = 1` as additional arguments in the `aes()` function. The `after_stat(prop)` is a statistical function used to calculate proportions. The `group = 1` implies that all the observations belong to one single group when calculating the proportions. We will try to set `group` to another variable in the next part.

```{r}
ggplot(data = sahp) + 
  geom_bar(mapping = aes(x = kit_qual, 
                         y = after_stat(prop), 
                         group = 1))
```




In addition to relying on `geom_bar()` to compute the frequencies or the proportions, we can also manually use the frequencies or proportions if we have the information in a data frame. Let's first create a data frame that contains the frequencies and proportions of each `kit_qual` value.

```{r, results = TRUE}
freq <- table(sahp$kit_qual)
prop <- freq/nrow(sahp)
kit_qual_stat <- data.frame(kit_qual = names(freq), 
                       freq = as.vector(freq), 
                       prop = as.vector(prop))
kit_qual_stat
```

Now, you can directly use the data frame `kit_qual_stat` to generate the bar chart. Here, you need to set `x = kit_qual` and `y = freq` in `aes()`, and `stat = "identity"` as an argument in the `geom_bar()` function. To use proportions instead of frequencies, you can use `y = prop` instead of `y = freq`.

```{r, results = TRUE}
ggplot(kit_qual_stat) + 
  geom_bar(aes(x = kit_qual, y = freq), 
           stat = "identity")
ggplot(kit_qual_stat) + 
  geom_bar(aes(x = kit_qual, y = prop), 
           stat = "identity")
```

You can easily check that the plots are identical to the previous bar charts. 

### Reordering Bars in Bar Charts {#reordering-bar-chart}
In our bar chart example, the bars are ordered alphabetically by default. Sometimes, we may want to reorder the bars according to certain criterion. 

***a. Reorder in ascending/descending order of heights***

To order the bars in ascending/descending order of their heights, you can use the `fct_reorder()`  function in the **forcats** package to reorder the factor. 

```{r}
library(forcats)
ggplot(data = sahp) + 
  geom_bar(mapping = aes(x = fct_reorder(kit_qual, 
                                         kit_qual, 
                                         length))) #increasing order
```

The above code reorders the bar chart to the increasing order of their heights. The `fct_reorder()` is a very powerful function used to reorder the levels of a factor according to a certain criterion. The function `fct_reorder(.f, .x, .fun = median)` has three arguments:

- `.f`: the discrete variable/factor to reorder
- `.x`: one variable
- `.fun`: the function to be applied to `.x`

The levels of `f` are reordered so that the values of `.fun(.x)`  are in ascending order. In our example, we reorder the levels of `kit_qual` such that the length of `kit_qual` is in ascending order, i.e. the bars are in ascending order of their heights. 

To reorder in descending order of heights, you can add an additional argument `.desc = TRUE` in the `fct_reorder()` function. 
```{r}
ggplot(data = sahp) + 
  geom_bar(mapping = aes(x = fct_reorder(kit_qual, 
                                         kit_qual, 
                                         length, 
                                         .desc = TRUE))) #decreasing order
```

***b. Manual Reorder***

In addition to reordering according to certain function values, you can use the function `fct_relevel()` to **manually reorder** the bars in a bar chart. In our example of `kit_qual`, perhaps a common thought is to order the levels from the worst quality to the best quality. 
```{r}
ggplot(data = sahp) +
  geom_bar(mapping = aes(x = fct_relevel(kit_qual, 
                                         c("Fair", 
                                           "Average", 
                                           "Good", 
                                           "Excellent"))))
```

As you can imagine, the second argument of `fct_relevel()` contains the desired order of the factors, which will be reflected in the order of the bars. 

### Aesthetics in Bar Charts {#bar-aes}

As before, we can use aesthetics to control the appearance of bar charts. 

First, let's look at a new aesthetic called **fill**, which fills the bar with different colors according to the value of the mapped variable (usually another discrete variable). 

::: {.infobox .caution data-latex="{caution}"}
From this section on, we will not separately discuss the constant-valued aesthetics and aesthetics mappings. If necessary, you are welcome to review the previous sections on aesthetics for scatterplots, line plots, and smoothline fits. 
:::

Here, we want to look at the distribution of `kit_qual` for different values of `central_air`.


```{r}
ggplot(data = sahp) + 
  geom_bar(mapping = aes(x = kit_qual, 
                         fill = central_air))
```



We can see that each bar is divided into stacked sub-bars with different colors. The different colors in each sub-bar correspond to the value of `central_air`. And the height of each sub-bar represents the count for the cases with a particular value of `kit_qual` and another value of `central_air`.

Let's verify the first bar.
```{r, results=TRUE}
sum(sahp$kit_qual == "Average" & sahp$central_air == "Y")
```

Clear, the result matches the blue portion of the first bar. 


When we map a variable to the `fill` aesthetic, the appearance of different subbars can be customized using the `position` argument as a constant-valued aesthetic in `geom_bar()`. 

***a. Stacked Bars***

The default value of the `position` argument is `"stack"`, which generated a collection of stacked bars with different colors. 

```{r, eval = FALSE}
ggplot(data = sahp) + 
  geom_bar(mapping = aes(x = kit_qual, 
                         fill = central_air), 
           position = "stack")
```


***b. Dodged Bars***

Using the stacked bars, it is sometimes difficult to compare the counts for different sub-bars. To make the comparison easier, you can use `position = "dodge"`.

```{r}
ggplot(data = sahp) + 
  geom_bar(mapping = aes(x = kit_qual, 
                         fill = central_air), 
           position = "dodge")
```

Now, the plot places the sub-bars beside one another, which makes it easier to compare individual counts for each combination of `kit_qual` and `central_air`.

***c. Filled Bars***

```{r}
ggplot(data = sahp) + 
  geom_bar(mapping = aes(x = kit_qual, fill = central_air), 
           position = "fill")
```

`position = "fill"` works like stacking, but makes each set of stacked bars the same height. Perhaps the `y` axis should be labeled as "proportion" rather than "count". This makes it easier to compare proportions of different values of `central_air` for different values of `kit_qual`. For example, we can see that the proportion of `central_air = TRUE` is much higher for `kit_qual = "Good"`  than that for `kit_qual = "Fair"`. 


Lastly, you can also use the polar coordinates. 
```{r}
ggplot(data = sahp) + 
  geom_bar(mapping = aes(x = kit_qual)) +
  coord_polar()
```

### Pie Charts {#pie-charts}

It turns out **pie charts** can be generated in a similar fashion as bar charts, by adding an argument in the `geom_bar()` function. Let's try to generate a pie chart for `kit_qual`. The idea is to first generate a single stacked bar with each sub-bar corresponding to different `kit_qual` values.

```{r}
ggplot(kit_qual_stat) +
  geom_bar(aes(x = "", 
               y = prop, 
               fill = kit_qual),
           stat = "identity")
```

Then, to get a pie chart, you just need to add an additional layer to change the coordinates from Cartesian coordinates into polar coordinates using the `coord_polar()` function.

```{r}
ggplot(kit_qual_stat, 
       aes(x = "", 
           y = prop, 
           fill = kit_qual)) +
  geom_bar(stat = "identity")  +
  coord_polar("y") +
  theme_void()
```

Sometimes, you may want to add texts on to the pie chart to represent the percentages of each category. You can do that by using the `geom_text()` function with `label` and `position` arguments specified.

```{r}
ggplot(kit_qual_stat, 
       aes(x = "", 
           y = prop,
           fill = kit_qual)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(100*prop, 2)),
            position = position_stack(vjust = 0.5)) +
  coord_polar("y") +
  theme_void()
```

### Exercises

Use the `sahp` data set to answer the following questions.

1. Create a bar chart to represent the distribution of the number of available car spaces in the garage (`gar_car`).

2. For the bar chart in Q1, divide each bar into sub-bars according to whether `oa_qual > 5`. What findings do you have in this plot? 

3. For the bar chart in Q2, change the position of the sub-bar to reflect the proportion for `oa_qual > 5` for each value of `gar_car`. 

4. Create a pie chart to represent the distribution of the number of available car spaces in the garage (`gar_car`) with the corresponding percentages displays inside the pie chart.
