## Customization in ggplot {#customization}


In this section, we would like to digress a little bit to show some possible 
customizations on the axes, labels, and titles. 

Let's first review the following scatterplot introduced in Section \@ref(scatterplot).

```{r, results = TRUE}
library(ggplot2)
library(r02pro)
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol))
```

Before introducing customization, let's introduce a way to save the current plot and add additional layers, to avoid repetition of codes. For example, we can break down the previous code into the following two steps.

```{r}
step_1 <- ggplot(data = gm2004)
step_1 + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol))
```

This is equivalent to the previous plot. Let's save the whole plot into an R object.

```{r}
my_plot <- ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol))
```


### Customization of Labels and Titles

***a. Customize x and y labels and title***

By default, the x and y labels are the variables names, and there is no title for the plot. You can customize the x and y labels using the `xlab()` and `ylab()` function, and add a title with the `ggtitle()` function. 

```{r, warning=FALSE}
my_plot_2 <- my_plot + 
  xlab("Sugar (grams per person and day)") + 
  ylab("Cholesterol (mmol/L)") + 
  ggtitle("Cholesterol vs. Sugar")
my_plot_2
```

***b. Customize the font of the x and y breaks***

In addition, you can further customize the font of the x and y breaks using the `theme()` function with the argument `axis.text`.  

```{r}
my_plot_2 + theme(axis.text = element_text(size = 25, color = "red"))
```


***c. Customize the font of labels***

To customize the font, you can use `axis.title` argument to change the size, color, and face of the labels. 
```{r}
my_plot_2 + theme(axis.title = element_text(size = 18, 
                                     color = "red", 
                                     face = "italic"))
```



***d. Customize the font of the title***

Similarly, you can use the `plot.title` argument to customize the font of the title. 

```{r}
my_plot_2 + theme(plot.title = element_text(size = 24, 
                                     color = "magenta", 
                                     face = "bold"))
```


***e. Center the title***

Sometimes, we may want to center the title. We can achieve this by setting the `hjust` parameter. 

```{r}
my_plot_2 + theme(plot.title = element_text(size = 24, 
                                     color = "magenta", 
                                     face = "bold",
                                     hjust = 0.5)) 
```


***f. Mix***

Apparently, you are free to mix all the different customizations. Let's see an example as below.
```{r}
my_plot_2 + theme(axis.title = element_text(size = 18, 
                                     color = "red", 
                                     face = "italic"), 
           axis.text = element_text(size = 12, 
                                    color = "blue"), 
           plot.title = element_text(size = 24, 
                                     color = "magenta", 
                                     face = "bold", 
                                     hjust = 0.5))
```

***g. Save as a theme***

As you can see from the code, the code gets complicated if we want to customize many things at the same time. To save time, you can actually save the desired into an R object and reuse it later. 
```{r}
mytheme <- theme(axis.title = element_text(size = 18, 
                                           color = "red", 
                                           face = "italic"), 
                 axis.text = element_text(size = 12, 
                                          color = "blue"), 
                 plot.title = element_text(size = 24, 
                                           color = "magenta", 
                                           face = "bold", 
                                           hjust = 0.5))
```

Then, we can generate the same plot with `mytheme` using

```{r, eval=FALSE}
g1 + mytheme
```
For a different plot, we can also use the same `mytheme`.

```{r}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = log10(GDP_per_capita), 
                           y = life_expectancy)) +
  mytheme
```


### Customizing the Breaks on the x and y Axes

In the above plot, the **breaks** on the x axis are 0, 50, 100, 150, and 200. The breaks on the y axis are 4.0, 4.5, 5.0, and 5.5. Sometimes, we may want to customize the breaks, e.g. to show a finer scale. To do this, we can use the `scale_x_continous()` and `scale_y_continuous()` functions. Both functions take an argument called `breaks` which is a numeric vector specifying the desired breaks on the x and y axes. 

```{r}
my_plot +
  scale_x_continuous(breaks = seq(from = 0, to = 200, by = 25)) +
  scale_y_continuous(breaks = seq(from = 4, to = 5.5, by = 0.25))
```

In this example, we have customized the breaks for `sugar` to be a equally-spaced sequence from 0 to 200 with increment 25, and the breaks for `cholesterol` to be another equally-spaced sequence from 4 to 5.5 with increment 0.25.

When the specified breaks do not cover the full range of the data, you will see the breaks changed but all the data points are still visible. 

```{r}
my_plot +
  scale_x_continuous(breaks = seq(from = 0, to = 150, by = 25)) +
  scale_y_continuous(breaks = seq(from = 4, to = 4.5, by = 0.25))
```

On the other hand, if the specified breaks go beyond the value of the data, `ggplot()` will only show the breaks values within the data range.

```{r}
my_plot +
  scale_x_continuous(breaks = seq(from = 0, to = 300, by = 25)) +
  scale_y_continuous(breaks = seq(from = 4, to = 8, by = 0.25))
```


### Zoom In to a Specific Region of the Data

Sometimes, you want to zoom in to the specific region of the data to see a finer detail. To do this, you can use the `coord_cartesian()` function with arguments `xlim` and `ylim` for specifying the desired region. 

Let's say we want to focus on the houses with `sugar` between 25 and 75. 

```{r}
my_plot +
  coord_cartesian(xlim = c(25, 75))
```

Let's narrow down further to `sugar` between 25 and 75 and `cholesterol` between 4.5 to 5.0.

```{r}
my_plot +
  coord_cartesian(xlim = c(25, 75),
                  ylim = c(4.5, 5))
```

### Generate Log-Scale Plot

Recall that we generated the following scatterplot between `GDP_per_capita` and `life_expectancy`, where we took a logarithm transformation of `GDP_per_capita`. 

```{r}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = log10(GDP_per_capita), 
                           y = life_expectancy))
```

For this particular `log()` transformation, an alternative is to generate a **Log-scale Plot** by setting the `trans` argument in the `scale_y_continous()` function. The `log-scale plot` is a popular way for displaying numerical data over a very wide range of values in a compact fashion. 

The default value of `trans` is `"identity"`, meaning no transformation. There are many different `trans` choices including `log`, `exp`, `log10`, `sqrt`, and others.

```{r}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = GDP_per_capita, 
                           y = life_expectancy)) +
  scale_x_continuous(trans = "log10")
```

Although the two figures look visually identical, the log-scale plot may be more intuitive to interpret than the plot on the transformed variable on the x-axis. 


### Coordinate Flip

In some situations, you may want to flip the x and y coordinates. To do this, you just need to add `coord_flip()` to the existing ggplot. In a future section, we will see other type of transformation for coordinates. 

```{r}
my_plot + coord_flip()
```


### Exercises

Use the `sahp` data set to answer the following questions.

1. Create a scatterplot between `lot_area` (x-axis) and `sale_price` (y-axis), with the breaks on the x-axis being an equally-spaced sequence from 0 to 40000 with increment 5000, and the breaks on the y-axis being (0, 200, 300, 550). And set the labels on the axes to be "Lot Area" and "Sale Price", and the title to be "Sale Price vs. Lot Area". Then, apply `mytheme` to the plot.

2. For the plot in Q1, create a zoom-in plot where `lot_area` is between 10000 and 15000, and `sale_price` is between 200 and 300. 

3. For the plot in Q1, create a corresponding log-log plot, where both the x-axis and y-axis are in log-scale. 






