## Histograms {#histogram}

In Section \@ref(bar-charts), we learned how to use `geom_bar()` to generate bar charts for visualizing the distributions of discrete variables. You may be wondering, how about visualizing continuous variables? One popular plot is called **histograms**. 

In this section, we will return to the `gm2004` data set.

### Using the `hist()` function

To generate a histogram, you can simply use `hist()` with the variable as the argument.  

```{r}
library(r02pro)
hist(gm2004$cholesterol)
```

On the x-axis, the histogram displays the range of values for the cholesterol value. Then, the histogram divides the x-axis into bins with equal width, and a bar is erected over the bin with the y-axis showing the corresponding number of observations (called Frequency on the y-axis label).  

### Using the `geom_histogram()` function

In addition to using the `hist()` function in base R. The `geom_histogram()` function in the **ggplot2** package provides richer functionality. Let's take a quick look. 

```{r}
library(ggplot2)
ggplot(data = gm2004) + 
  geom_histogram(mapping = aes(x = cholesterol))
```

We could see a message, saying "`stat_bin()` using `bins = 30`" which implies the histogram has 30 bins by default. Next, we introduce three different ways to customize the bins. 

***a. Use Aesthetic `bins`.***

We can change the number of bins with the constant-valued aesthetic `bins`.

```{r, warning=FALSE}
ggplot(data = gm2004) + 
  geom_histogram(mapping = aes(x = cholesterol),
                 bins = 5)
```

We can see that the histogram now has 5 bins and each bin has the same width. 

***b. Use Aesthetic `binwidth`.***

And from the message, another way to change the number of bins is to specify the `binwidth`, which is another constant-valued aesthetic.

```{r, warning=FALSE}
ggplot(data = gm2004) + 
  geom_histogram(mapping = aes(x = cholesterol),
                 binwidth = 0.2)
```

***c. Manually Set the Bins.***

If desired, you can manually set the bins via the `breaks` argument in the `geom_histogram()` function. 

```{r, warning=FALSE}
ggplot(data = gm2004) + 
  geom_histogram(mapping = aes(x = cholesterol),
                 breaks = seq(from = 3.5, 
                              to = 6, 
                              by = 0.5))
```

The `breaks` argument is a numeric vector specifying how the bins are constructed. Let's verify the height of the first bin. 

```{r}
#verify the first bin count
sum(gm2004$cholesterol < 4, na.rm = TRUE)
```

The number matches to the height of the first bin. Not that we need to add `na.rm = TRUE` as an argument since there are missing values.

### Aesthetics in `geom_histogram()`

Next, we introduce the aesthetics in histograms, which are very similar to those in bar charts. For example, we can map a variable to the fill aesthetic.

```{r, warning=FALSE}
ggplot(data = gm2004) + 
  geom_histogram(mapping = aes(x = cholesterol,
                               fill = gender),
                 bins = 5)
```

We can see that like the bar chart, the bar for each bin is now divided into sub-bars with different colors. The different colors in each sub-bar correspond to the values of `gender`. And the height of each sub-bar represents the count for the cases with the `cholesterol` in this particular bin and the specific value of  `gender`. We observe that while the male and female group behave similarly in the middle bins, many more countries has a small mean cholesterol value for males than those for females.

Just like `geom_bar()`, we have a global aesthetic called `position`, which does position adjustment for different sub-bars. The default position value is again `"stack"` if you don't specify it. 

***a. Stacked Bars***

```{r, eval=FALSE}
ggplot(data = gm2004) + 
  geom_histogram(mapping = aes(x = cholesterol,
                               fill = gender),
                 position = "stack",
                 bins = 5)
```

As expected, we get the same histogram as before.

***b. Dodged Bars***

The second option for position is `"dodge"`, which places the sub-bars beside one another, making it easier to compare individual counts for the combination of a bin of `cholesterol` and `gender`.

```{r, warning=FALSE}
ggplot(data = gm2004) + 
  geom_histogram(mapping = aes(x = cholesterol,
                               fill = gender),
                 position = "dodge",
                 bins = 5)
```


***c. Filled Bars***

Another option for optional is `position = "fill"`. It works like stacking, but makes each set of stacked bars the same height. 

```{r, warning=FALSE}
ggplot(data = gm2004) + 
  geom_histogram(mapping = aes(x = cholesterol,
                               fill = gender),
                 position = "fill",
                 breaks = seq(from = 3.5, 
                              to = 6, 
                              by = 0.5))
```

Just like `geom_bar()`, the y axis may be labeled as "proportion" rather than "count", to be more precise. This makes it easier to compare proportions of different values of gender for different bins of cholesterol. For example, we can see that the proportion of `gender = "male"` is  100% for cholesterol values smaller than 4.  

### Density Estimate Using Histograms

In addition to using histograms to visualize the distribution of a discrete variable, you can also construct a density density of variable using a proper normalization. To generate such a density estimate, you can add `y = ..density..` as a mapping in the `aes()` function in `geom_historgram()`. Let's see an example as below.
```{r, warning=FALSE}
ggplot(data = gm2004) + 
  geom_histogram(mapping = aes(x = cholesterol,
                               y = ..density..),
                 breaks = seq(from = 3.5, 
                              to = 6, 
                              by = 0.5))
```
Let's try to calculate the height of the first bar together. We know that the total area of the bars is 1, agreeing with the definition of density. First, we get the area of the first bar. Then we divide it by the width to get the height.  

```{r, results=TRUE}
cholesterol_no_na <- na.omit(gm2004$cholesterol) 
sum(cholesterol_no_na < 4)/length(cholesterol_no_na) #area of the first bar
sum(cholesterol_no_na < 4)/length(cholesterol_no_na) /0.5 #height of the first bar
```

You can see that the height matches the `y` axis for the first bar.

### Exercises
Use the `sahp` data set to answer the following questions.

1. Create three different histograms on the living area (`liv_area`) for each of the following settings

- Use 10 bins
- Set the binwidth to be 300
- Se the bins manually to an equally-spaced sequence from 0 to 3500 with increment 500. 

2. Create  histograms on the living area (`liv_area`) with 5 bins, and show the information of different `kit_qual` values in each bar. What conclusions can you draw from this plot? 



