## ggplot Aesthetics (III): Global Aesthetic Mapping {#multi-geom}

So far, you have learned how to create scatterplots using `geom_point()`, line plots using `geom_line()`, and smoothline fits using `geom_smooth()`. It is sometimes useful to combine multiple geoms in the same plot. 

Let's first review the scatterplot and smoothline fit between `sugar` and `cholesterol`. 

```{r, warning=F, message=F}
library(r02pro)
library(tidyverse)
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol))
ggplot(data = gm2004) + 
  geom_smooth(mapping = aes(x = sugar, 
                           y = cholesterol))
```

### Multiple geoms

To combine multiple geoms, you can simply use `+` to add them. 
```{r, warning=F}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol)) + 
  geom_smooth(mapping = aes(x = sugar, 
                           y = cholesterol))
```

As expected, you see all the points and the smoothline fit on the same plot, which contains very rich information.


As usual, you can add aesthetics to both geoms. 

Let's first map `gender` to the color aesthetic for `geom_smooth()`.
```{r, warning=F}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol)) + 
  geom_smooth(mapping = aes(x = sugar, 
                           y = cholesterol,
                           color = gender))
```

To verify the two smoothline fits are indeed fitted from the data points in the two groups, you can map `gender` to the color aesthetic for `geom_point()` as well. In addition, let's also map `population` to the size aesthetic. 
```{r, warning=F}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           color = gender,
                           size = population)) + 
  geom_smooth(mapping = aes(x = sugar, 
                           y = cholesterol,
                           color = gender))
```

The plot is reassuring that the two smoothline fits indeed correspond to the data points in the two groups defined by `gender`. 


In addition to mapping variables to aesthetics, you can also add Constant-Valued aesthetics for each geom.
```{r, warning=F}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol,
                           color = gender,
                           size = population),
             shape = 2,
             alpha = 0.9) + 
  geom_smooth(mapping = aes(x = sugar, 
                           y = cholesterol,
                           color = gender),
              linetype = 2,
              se = FALSE)
```

Now, let's see an example of combining the scatterplot with the line plot using the `sahp` dataset. Let's focus on the houses that were sold before 2007. 
```{r}
sahp_2006 <- sahp[format(sahp$dt_sold, "%Y") < 2007, ] #all houses sold before 2007
ggplot(data = sahp_2006) + 
  geom_line(mapping = aes(x = dt_sold, 
                          y = sale_price))
```

Next, we will add the line plot on top of the scatterplot. 
```{r}
ggplot(data = sahp_2006) + 
  geom_line(mapping = aes(x = dt_sold, 
                          y = sale_price)) + 
  geom_point(mapping = aes(x = dt_sold, 
                          y = sale_price))
```

Clearly, having both the scatterplot and the line plot make the plot more informative. 

### Global Aesthetic Mapping {#global-aes-map}

In Sections \@ref(aes-constant-value) and \@ref(aes-local-map), we introduced the Constant-Valued Aesthetics and Local Aesthetic Mapping. Now, we are going to introduce the final way of specifying aesthetics, namely the **Global Aesthetic Mapping**. In contrast to the local aesthetic mapping where the mapping only works at the local level, the global aesthetic mapping will impact all the geoms considered in the entire ggplot. 

Let's take another look at the example where we have a scatterplot and a smoothline fit for the relationship between `cholesterol` and `sugar`. 

```{r, eval=FALSE}
ggplot(data = gm2004) + 
  geom_point(mapping = aes(x = sugar, 
                           y = cholesterol)) + 
  geom_smooth(mapping = aes(x = sugar, 
                           y = cholesterol))
```


You may notice that the arguments inside `geom_point()` and `geom_smooth()` are identical. The global aesthetic mapping allows us to simplify codes by specifying the mapping only once at the `ggplot()` level. Let's take a look at how it works.

```{r, eval=FALSE}
ggplot(data = gm2004,
       mapping = aes(x = sugar,
                     y = cholesterol)) + 
  geom_point() + 
  geom_smooth()
```

Clearly, this code is simpler to read and avoid repetition. 

As another example, for the combination of scatterplot and line plot, we can also take advantage of the global aesthetic mapping.

```{r, eval = FALSE}
ggplot(data = sahp_2006,
       mapping = aes(x = dt_sold, 
                     y = sale_price)) + 
  geom_line() + 
  geom_point()
```


Naturally, we can add additional aesthetic mappings to the `aes()` function.

```{r, warning=FALSE}
ggplot(data = gm2004,
       mapping = aes(x = sugar,
                     y = cholesterol,
                     color = gender)) + 
  geom_point() + 
  geom_smooth()
```

#### Extend Global Aesthetic Mappings

In addition to the global aesthetic mappings, you can add additional local aesthetic mappings to each geom.

```{r, warning=FALSE}
ggplot(data = gm2004,
       mapping = aes(x = sugar,
                     y = cholesterol,
                     color = gender)) + 
  geom_point(mapping = aes(size = population)) + 
  geom_smooth()
```


Here, the global aesthetic mapping `x = sugar`, `y = cholesterol`, and `color = gender` are passed to `geom_point()` and `geom_smooth()`. And we have an additional local aesthetic mapping `size = population` for `geom_point()`. 

#### Overwrite Global Aesthetic Mappings

When you add the same aesthetic mapping in the local geom as one of the global mappings, the corresponding global aesthetic mapping will be **overwritten**. 

```{r, warning=FALSE}
ggplot(data = gm2004,
       mapping = aes(x = sugar,
                     y = cholesterol,
                     color = gender)) + 
  geom_point(mapping = aes(color = continent)) + 
  geom_smooth()
```

Here, the global aesthetic mapping `color = gender` is first passed to both `geom_point()` and `geom_smooth()`. However, since there is a local aesthetic mapping for color (`color = continent`) in `geom_point()`, the local aesthetic mapping will **overwrite** the global one, resulting countries in different continents having different colors. Note that the legend may look a bit weird as it contains a mix of values for continent and gender under the name gender. This is an artifact of the `ggplot` implementation. As a result, it is recommended to avoid this kind of overwriting behavior. 

#### No Global Constant-Valued Aesthetics

Let's review the difference between the Constant-Valued Aesthetic and the Global Aesthetic Mapping. 

- The constant-valued aesthetic set the aesthetic value for all points/lines on the graph (for example, `color = red` will make all points red) for a single geom. It doesn't interfere with the other geoms. 
- The global aesthetic mapping is an aesthetic mapping (for example, `color = gender` will use different colors for different gender values) that will be passed into all available geoms considered in the single plot. 

It is worth noting that we can't set the constant-valued aesthetic at the global level, i.e.there is no **Global Constant-Valued Aesthetics**. Let's try to change all the points and lines to purple by adding `color = "purple"` in `ggplot()`. 

```{r, warning=FALSE}
ggplot(data = gm2004,
       mapping = aes(x = sugar,
                     y = cholesterol),
       color = "purple") + 
  geom_point() + 
  geom_smooth()
```

As you can see from the output, the argument `color = "purple"` was ignored during the plotting process. To achieve this, you need to use `color = "purple"` as constant-valued aesthetics for both geoms as below.

```{r, warning=FALSE}
ggplot(data = gm2004,
       mapping = aes(x = sugar,
                     y = cholesterol)) + 
  geom_point(color = "purple") + 
  geom_smooth(color = "purple")
```

#### Mixing the three types of aesthetics

Now, let's see an example where all three types of aesthetics are used.

```{r, warning=FALSE}
ggplot(data = gm2004,
       mapping = aes(x = sugar,
                     y = cholesterol)) + 
  geom_point(mapping = aes(size = population),
             color = "magenta") + 
  geom_smooth(mapping = aes(color = gender),
              se = FALSE)
```


### Exercises

Using the `sahp` dataset in the **r02pro** package, using the **ggplot2** package to answer the following questions.

1. With `lot_area` on the x-axis and `sale_price` on the y-axis, create a plot that contains both the scatterplot and smoothline fits, where we use different colors in the scatterplot to distinguish whether `heat_qual` is excellent and different linetypes for the smoothline fits depending on whether `house_style` is 2Story.
2. Create a plot of `liv_area` (on the x-axis) and `sale_price` (on the y-axis) that contains both the scatterplot and the smoothline fit.
3. In the plot from Q2, using different colors for both the scatterplot and the smoothline fit to distinguish whether the house has more than 3 bedrooms and make all points to be size 2.
4. If you run the following code 
```{r, eval=FALSE}
ggplot(data = sahp, 
       mapping = aes(x = liv_area, 
                     y = sale_price), 
       color = "green") + 
  geom_point() + 
  geom_smooth()
```

Do you think all the points and the smoothline will be green? If not, explain the reason and make them green. 

5. Run 
```{r, eval = FALSE}
ggplot(data = sahp, 
       mapping = aes(x = liv_area, 
                     y = sale_price, 
                     color = house_style)) + 
  geom_point(mapping = aes(color = bedroom > 3)) +
  geom_smooth(mapping = aes(color = bedroom > 3))
```

Explain why you only see two colors in the plot although the `house_style` is mapped to color? 




