## Error Bars {#errorbar}

So far, we have learned several ways to compare continuous distributions for different groups, including the density plot, boxplot, and the violin plot. In this lesson, we introduce another visualization method, called **error bar**, which is a graphical representations of the uncertainty in a certain measurement. Let's prepare a ggplot object to get started.

```{r, warning=FALSE}
library(r02pro)
library(ggplot2)
g <- ggplot(data = sahp, 
            aes(x = house_style, 
                y = sale_price)) + 
  scale_x_discrete(limits=c("1Story", 
                            "2Story"))
```

Here, we visualize the error bar as the mean $\pm$ standard error. 

To show the mean of `sale_price` for different `house_style` groups, you can use the `geom_point()` function with arguments `stat = "summary"` and `fun = "mean"`. 

```{r, warning=FALSE}
g  + geom_point(stat = "summary", 
                fun = "mean")
```

To add error bars to the mean points, we use `geom_errorbar()` coupled with the `fun.data = "mean_se"` argument.

```{r, warning=FALSE}
g  + geom_point(stat = "summary", 
                fun = "mean") + 
  geom_errorbar(stat = "summary", 
                fun.data = "mean_se") 
```

You can control the width of the errorbar by setting the `width` argument. 
```{r, warning=FALSE}
g + geom_point(stat = "summary", 
                fun = "mean") + 
  geom_errorbar(stat = "summary", 
                fun.data = "mean_se", 
                width = 0.2)
```

This plot is called a **mean-error-bar** plot. Let's take a detailed look at this plot. The distance from the mean point to the lower bound of the error bar is the so called **standard error**, which is defined as the standard deviation of the sample divided by the square root of the sample size. Let's try to calculate the upper and lower bounds of the error bar. 

```{r, results=TRUE}
sale_price_1_story <- sahp$sale_price[sahp$house_style == "1Story"]
sale_price_1_story_se <- sd(sale_price_1_story)/sqrt(length(sale_price_1_story))
mean(sale_price_1_story) + sale_price_1_story_se
mean(sale_price_1_story) - sale_price_1_story_se
```

You may notice that we are referring the data as sample since the `sahp` dataset contains only a subset of the population with all houses in the world. An important application of error bar is to construct the 95% confidence interval of the population mean. You know that the 95% confidence interval for a normal distribution is mean $\pm$ 1.96*se. If the data follows a normal distribution, you can construct such a confidence interval by setting the multiplier of the standard error in the error bar to be 1.96.  

```{r, warning=FALSE}
g + geom_point(stat = "summary", 
                fun = "mean") + 
  geom_errorbar(stat = "summary", 
                fun.data = "mean_se", 
                width = 0.2, 
                fun.args = list(mult = 1.96))
```

Of course, you can also add the error bar on top of the other plots like the boxplot or violin plot.
```{r, warning=FALSE}
g  + geom_point(stat = "summary", 
                fun = "mean") + 
  geom_boxplot() +
  geom_errorbar(stat = "summary", 
                fun.data = "mean_se", 
                width = 0.2)
```

```{r}
g  + geom_point(stat = "summary", 
                fun = "mean") + 
  geom_violin() + 
  geom_errorbar(stat = "summary", 
                fun.data = "mean_se", 
                width = 0.2)
```

