## Heat Map {#heat-map}

A **heat map** is a data visualization technique that shows magnitude of a phenomenon in the two dimensional space. 

First of all, we prepare a smaller data set from the `gm2004` data in the **r02pro** package, to better illustrate the mechanism of a heat map.

```{r}
library(r02pro)
library(tidyverse)
country_list <- c("Australia", 
                  "China",
                  "United States",
                  "United Kingdom",
                  "South Africa")
sgm2004 <- gm2004 %>%
  dplyr::filter(gender == "female", country %in% country_list) %>%
  select(country, population, BMI, cholesterol, GDP_per_capita)
```

Note that during the data preparation process, we have used the `filter()` and `select()` functions, which filters observations and select variables from a tibble, respectively. They will be covered in detailed in the next Chapter. 


### Using the `heatmap()` function

We will first introduce the function `heatmap()`, which is available in base R. To use the function, you need to convert the object into a `matrix` using `as.matrix()` and specify the rownames of the object. 

```{r}
sgm2004_mat <- as.matrix(sgm2004[, -1])
rownames(sgm2004_mat) <- country_list
sgm2004_mat
heatmap(sgm2004_mat)
```

The working of mechanism of a heatmap is that it creates a grid of colored gray-scale rectangles with colors corresponding to the values in the corresponding matrix. By default, the **rows** of the input matrix are scaled to have mean zero and standard deviation one. In our example, perhaps it is more meaningful to scale the **columns** as the represent different variables. We can achieve this by setting `scale = "column"`. 

```{r}
heatmap(sgm2004_mat, scale = "column")
```

In addition to generating the heat map, the `heatmap()` function, by default, also run a **clustering algorithm** on both the rows and columns, and visualize the results with **dendrograms**. To turn off the dendrograms, you can set arguments `Rowv = NA` and `Colv = NA` for the row and column dendrograms, respectively. 

In addition, it is easy to customize the colors of the heatmap.

```{r}
rc <- rainbow(nrow(sgm2004_mat), start = 0, end = .3)
cc <- rainbow(ncol(sgm2004_mat), start = 0, end = .3)
heatmap(sgm2004_mat, 
        col = cm.colors(256),
        scale = "column",
        RowSideColors = rc, 
        ColSideColors = cc)
```

### Using `geom_tile()` in **ggplot2**

In the **ggplot2** packages, there is no geom that directly generates a heatmap, however, we can use the `geom_tile()` on a transformed data format using `pivot_longer()`, which will be introduced in Chapter ??. 

```{r}
sgm2004_long <- sgm2004 %>%
  pivot_longer(cols = 2:5, 
               names_to = "Var")
ggplot(sgm2004_long) +
  geom_tile(mapping = aes(x = country, 
                          y = Var,
                          fill = value))
```

Here, since the variables are not in the same scale, the colors of the tiles are dominated by the `population` value. As before, we need to scale the variables to make them comparable. 
```{r}
sgm2004_long <- sgm2004_long %>% 
  group_by(Var) %>%
  mutate(std_value = scale(value)) %>%
  ungroup()
ggplot(sgm2004_long) +
  geom_tile(mapping = aes(x = country, 
                          y = Var,
                          fill = std_value))
```

Now, the heatmap looks more informative that shows the relative magnitude across different countries for each variable considered. 

Like in Section \@ref(aes-local-map), you can customize the range of the tile colors by specifying the `low` and `high` parameters in the `scale_fill_gradient()` function. 
```{r}
ggplot(sgm2004_long) +
  geom_tile(mapping = aes(x = country, 
                          y = Var,
                          fill = std_value)) + 
  scale_fill_gradient(low = "green", 
                      high = "red")
```

