## Line Segments {#add-auxiliary}

It is sometimes helpful to add line segments as auxiliary lines to existing plots to provide additional information. 

### Using `abline()` with `plot()`

Let's first review the scatterplot between `liv_area` and `sale_price`. 

```{r, warning=FALSE, fig.show='hide'}
library(r02pro)
plot(sahp$liv_area, sahp$sale_price)
```

In this plot, you may want to add some auxiliary lines. You can use the function `abline()` after a call of `plot()` to do this. To add a vertical line, you can set the parameter `v`; to add a horizontal line, you can set the parameter `h`.

```{r}
plot(sahp$liv_area, sahp$sale_price)
abline(v = 2000, col = "purple")       ##Add a vertical line at liv_area = 2000
```


```{r}
plot(sahp$liv_area, sahp$sale_price)
abline(h = 300, col = "blue")         ##Add a horizontal line at sale_price = 2000
```

Note that the `v` and `h` arguments can also be vectors with more than one values, which will lead to multiple vertical or horizontal lines.  The corresponding argument `col` can also be a vector with multiple values. 
```{r}
plot(sahp$liv_area, sahp$sale_price)
abline(h = c(100, 200, 300), 
       col = c("red","blue","green"))         ##Add multiple horizontal lines
```


In addition to adding vertical lines and horizontal lines, you can also add any line with the `abline()` function. We know a line can be represented as a function $y = a + b\times x$, where $a$ is the intercept and $b$ is the slope. In the `abline()`, you can generate such a line by specifying the parameter `a` for the intercept and `b` for the slope. Note that you can run  `abline()` multiple times to add multiple lines.

```{r}
plot(sahp$liv_area, sahp$sale_price)
abline(h = 300, col = "blue")   
abline(a = 100, b = 0.1, col = "green") 
```

### Using `geom_hline()`, `geom_vline()` and `geom_abline()`

Let's first review the following scatterplot between `liv_area` and `sale_price`. 

```{r, warning=FALSE}
library(r02pro)
library(tidyverse)
ggplot(data = sahp) + 
  geom_point(mapping = aes(x = liv_area, y = sale_price))
```

Looking at the scatterplot, it maybe helpful to add a horizontal line. To do this, you can use the `geom_hline()` function with argument `yintercept` specifying the value on the y-axis. 

```{r, warning=FALSE}
ggplot(data = sahp) + 
  geom_point(mapping = aes(x = liv_area, y = sale_price)) + 
  geom_hline(yintercept = 300, color = "red")
```

Here, a horizontal line at 300 is added to the scatterplot. 

You can also add both vertical lines and horizontal lines to the same plot. 

```{r, warning=FALSE}
ggplot(data = sahp) + 
  geom_point(mapping = aes(x = liv_area, y = sale_price)) + 
  geom_vline(xintercept = 2000, color = "green") + 
  geom_hline(yintercept = 300, color = "red")
```

In addition to adding vertical lines and horizontal lines, you can also add any line with the `geom_abline()` function. We know a line can be represented as a function $y = a + b\times x$, where $a$ is the intercept and $b$ is the slope. In the `geom_abline()`, you can generate such a line by specifying the `slope` and `intercept` arguments. 

```{r, warning=FALSE}
ggplot(data = sahp) + 
  geom_point(mapping = aes(x = liv_area, y = sale_price)) + 
  geom_abline(slope = 0.1, intercept = 100, color = "blue")
```

Similar to smoothline fit and line plots, you can change line type here. Different numbers correspond to different line types.

```{r}
ggplot(data = sahp) + 
  geom_point(mapping = aes(x = liv_area, y = sale_price)) + 
  geom_abline(slope = 0.1, intercept = 100, linetype = 3) 
```

### Exercises


1. Using the `sahp` data, using `plot()` to generate a scatterplot between `lot_area` (x-axis) and `sale_price` (y-axis), and add horizontal lines at 150 with color "blue" and 200 with color "red".
2. Using the `sahp` data, using `ggplot()` to generate a scatterplot between `sale_price` (x-axis) and `liv_area` (y-axis), and add the following lines to the plot: 
- $y = 5\times x + 1000$ (in green color, dashed line)
- $y = 3\times x + 1500$ (in purple color, solid line)





