## Interative Plot with `plotly()` {#plotly}

Until now, all the plots we created are all static. In this section, we will introduce a powerful tool named `plotly()` that can make interactive plots. If you haven't done so, you need to first install the R package **plotly**. 

```{r, eval = FALSE}
install.packages("plotly")
```

Let's start with a static scatterplot and make it interactive. 

```{r, warning=FALSE}
library(r02pro)
library(ggplot2)
library(plotly)
my_plot <- ggplot(data = gm2004) + 
    geom_point(mapping = aes(x = sugar, 
                             y = cholesterol))
ggplotly(my_plot)                           
```

Upon a first look, this may look identical to a regular scatterplot. However, you can try to move your cursor to the points which will show the corresponding coordinates. Some other features offered are available via a bar of buttons on the top right of the plot. Some useful features include 

- Download plot as png. 
- Zoom: Zoom a region of the plot.
- Pan: Move the plot around. 


In addition to the vanilla scatterplots, you can use plotly with more complicated plots that involves aesthetics. 

```{r}
my_plot_continent <- ggplot(data = gm2004) + 
    geom_point(mapping = aes(x = sugar, 
                             y = cholesterol,
                             color = continent))
ggplotly(my_plot_continent)   
```

Now, you can easily see the `continent` in addition to the `sugar`and `cholesterol` values for each data point. 

Let's try to use plotly with some other types of plots. 

```{r}
my_box_plot <- ggplot(data = na.omit(sahp), 
            aes(x = house_style, 
                y = sale_price)) +
  scale_x_discrete(limits=c("1Story", "2Story")) + geom_boxplot(mapping = aes(color = oa_qual > 5))
ggplotly(my_box_plot) 
```

For boxplots, we can easily see the values of summary statistics. You are welcome to try out plotly on other types of plots. 
