## Reorder Observations {#reorder-observations}

Now, let's look at the second task: find the 10 countries with the highest life expectancy in 2008. To order observations, you can use the function `arrange()` in the **dplyr** package. 

First, let's create a new dataset called `gm_2008` that only contains the observations in 2008. 

```{r}
library(r02pro)
library(dplyr)
library(tibble)
gm_2008 <- filter(gm, year == 2008)
```


To arrange the observations in the ascending order of the life expectancy (`life_expectancy`), you just need to add `life_expectancy` as a second argument of the `arrange()` function. To arrange the descending order, you can add `desc()` around the variable. 

```{r, results=TRUE}
arrange(gm_2008, yr_built)         #arrange in the ascending order of life_expectancy
arrange(gm_2008, desc(yr_built))   #arrange in the descending order of life_expectancy
```

You may observe from the results that there are several houses with the same `yr_built` value, leading to a tie. To break the tie, you can supply additional variables in the `arrange()` function, which will arrange the observations in the tie according to the additional variables sequentially.

```{r, results=TRUE}
arrange(jan09, desc(yr_built), bldg_class)
```

Here, the observations are arranged in the descending order of `yr_built`, and the ties are broken in the ascending order of `bldg_class`. Clearly, you can supply as many arguments as needed in the `arrange()` function. It is also important to note that the observations that has an `NA` value in the specified variable will always be arranged at the end of the output. 

### Exercises

Using the `ahp` dataset,

1. Find all houses built in `2008` with house style as `2Story`, then arrange the the observations in the ascending order of remodel year, and break the ties in the descending order of sale price.



