## Grouped Mutate and Filter via `group_by()` {#grouped-mutate-filter}

In Section \@ref(create-grouped-summaries), you  have seen the power of `group_by()` and `summarize()` which can help to create grouped summaries. In turns out the `group_by()` can be combined with `filter()` and `mutate()` to filter observations by group and create new variables by group. 

### Grouped Filter

For the `gm` dataset, let's say we want to find the countries that have a higher average GDP per capita than the average in the corresponding continent in the year 2008. We can use the following code.
```{r}
library(r02pro)
library(dplyr)
gm %>% 
  filter(year == 2008) %>%
  group_by(continent) %>%
  mutate(ave_GDP = mean(GDP_per_capita, na.rm = TRUE)) %>%
  filter(GDP_per_capita > ave_GDP) %>%
  select(country, continent, GDP_per_capita, ave_GDP)
```

The code works in the following steps.
1. Filter the observations for the year 2008.
2. Group the observations by `continent`.
3. Compute the average GDP per capita for each continent.
4. Filter the observations that have a GDP per capita higher than the average GDP per capita of the corresponding continent.
5. Select the variables `country`, `continent`, `GDP_per_capita`, and `ave_GDP`.

### Grouped Mutate

Now, let's say we want to find the top two countries with the highest GDP per capita for each continent in the year 2008. We can use the following code.
```{r}
gm %>% 
  filter(year == 2008 & !is.na(continent)) %>%
  group_by(continent) %>%
  mutate(r = rank(desc(GDP_per_capita), ties.method = "first")) %>%
  filter(r <= 2) %>%
  select(country, continent, GDP_per_capita) %>%
  arrange(continent, desc(GDP_per_capita))
```
Here, we first filter the observations for the year 2008 and remove the missing values in the `continent` variable. Then, we group the observations by `continent`. We use the `mutate()` function to create a new variable `r` which represents the rank of the GDP per capita in descending order. The `ties.method = "first"` argument is used to break the ties by giving the same rank to the same values. We then filter the observations that have a rank less than or equal to 2. Finally, we select the variables `country`, `continent`, and `GDP_per_capita`, and arrange the observations according to the continent and the GDP per capita in descending order.

### Exercises

Using the `ahp` dataset and the pipe operator for the following exercises.

1. For each month when the house was sold, find the houses that have a higher sale price than the average sale price in the corresponding month. Then, create a scatterplot between the month (x-axis) and the sale price (y-axis) with different colors for the houses that have a higher sale price than the average sale price in the corresponding month. Explain the findings from the figure.
2. Focus on the overall conditions where the number of corresponding houses are at least 30. Then, summarize the number of observations and the average sale price for each overall condition.
3. Find the top two expensive houses that was remodel each year, show only the remodel year and sale price, and arrange the result in the ascending order of remodel year. 
4. Group the houses by the combination of kitchen quality (`kit_qual`) and  central AC (`central_air`),  find the two houses with the minimum and maximum sale prices for each group. Then, arrange the groups in the descending order of maximum sale price within each group.


