## Select, Rename, and Reorder Variables {#select-variables}

In Section \@ref(filter-observations), we have learned how to filter observations according to a set of conditions. In many applications, there are some variables which may not be the focus of the research problem. As a result, learning how to **select** variables is very useful in data analysis. 

### Select Variables via Names

Let's consider the `gm` data set in the **r02pro** package, which contains `r ncol(gm)` variables. If we want to focus on the country, continent, year, and Human Development Index (HDI) category, we can use the `select()` function in the **dplyr** package. We will be using the pipe operator introduced in Section \@ref(pipe) throughout the rest of this Chapter. Inside the `select()` function, you can include the variables to select as different arguments. Note that the quotation marks are not necessary. 

```{r}
library(r02pro)
library(dplyr)
gm %>% 
  select(country, continent, year, HDI_category)
```

As you can see from the output, the `select()` function returns a tibble with only the selected variables.

Now, let's say we only want the observations for the year 2006. We can use the `filter()` function to get the subset of the `gm` data set with only the year 2006, and then use the `select()` function to keep only the variables of interest. 
```{r}
gm %>% 
  filter(year == 2006) %>% 
  select(country, continent, year, HDI_category)
```



### Rename Variables

In the `select()` function, you can also **rename the variables** in the form of `new_name = old_name` as the arguments. Let's see the following example where we want to rename the `country` to `nation` and `HDI_category` to `HDI_cat`. 
```{r}
gm %>% 
  select(nation = country, 
         continent, 
         year, 
         HDI_cat = HDI_category)
```


Note that in this renaming operations, only the variables mentioned in the `select()` function are kept. If you want to generate a tibble with some variables renamed while keeping all variables, you can use the `rename()` function.

```{r}
gm %>% 
  rename(nation = country, 
         HDI_cat = HDI_category)
```


From this example, you can see the different behavior between `select()` and `rename()` when renaming variables. 

### Select Variables with Patterns

Sometimes, you may want to select many variables that follows a particular pattern.

***a. select consecutive variables***

The first type of variable patterns is **consecutive variables** in a tibble.

To get started, you can use `select(1:m)` to get the first `m` variables. To get the last column, you can use `select(last_col())`. To get the last `m` variables, you can use `select(last_col(m-1):last_col())`.
```{r, results=TRUE}
gm %>% select(1:3)                    # the first 3 columns
gm %>% select(last_col())             # the last column
gm %>% select(last_col(1))            # the second to last column
gm %>% select(last_col(2):last_col()) # the last 3 columns
```


If you want to select the variable `country` and all variables from `HDI` to `continent` in the tibble, you can use `HDI:continent` as a single argument in the `select()` function. Note that the result highly depends on the particular ordering of variables in the tibble. 
```{r}
gm %>% 
  select(country, HDI:continent)
```


***b. select variables with names that start with, end with, or contain a string***

Another commonly used pattern is to select all variables with names that start with, end with, or contain a string. To select all variables that end with `"female"` (representing the female related measures) in the `gm` data set, you can use the `ends_with()` function in the `select()` function.

```{r}
gm %>% 
  select(ends_with("female"))
```

In the output, you can see the tibble with all variables ending with `"female"`. 

You can also use `starts_with()` to select all variables with names that end with a particular string. 
```{r, results=TRUE}
gm %>% 
  select(starts_with("BMI"))   #starts with BMI
```

This produces a tibble with all variables starting with `BMI`.

If you want to select all variables with names containing a string (not necessarily starts with or ends with), you can use the `contains()` function with the string as its argument.

```{r}
gm %>% 
  select(contains("newcases"))
```

This produces a tibble with all variables containing the string `"newcases"`. 

You can also combine multiple conditions using logical operators. First of all, the multiple arguments in the `select()` function are combined with or operations (`|`). For example, to get all variables that begin with `smoking` or end with `female`, you can use the following code.

```{r, results = TRUE}
gm %>% 
  select(starts_with("smoking"), ends_with("female"))
```

In the output, you can see the tibble with all variables starting with `"smoking"` or ending with `"female"`.


### Reorder Variables via Names

Knowing how to select and rename variables, let's learn how to reorder variables in a tibble. The first approach to reordering variables is to use the `select()` function by listing all the variables in the desired order. However, you may realized that it would be time consuming if there are many variables in the tibble and we only want to move a few to the beginning. It turns out we can utilize the `everything()` function as an argument in the `select()` function, which will select all the remaining variables. Let's say we want to move `country` and `continent` to the beginning of the tibble. 
```{r}
gm %>% 
  select(country, 
         continent, 
         everything())
```

This produces a tibble with `country` and `continent` as the first two columns, followed by the remaining variables in the same order as in the original tibble.


### Exercise

Using the `ahp` dataset and the pipe operator for the following exercises.

1. Select the variables `dt_sold`, `house_style`, `liv_area`, and `sale_price`, while renaming `dt_sold` to `date_sold` and `house_style` to `house_type`.

2. Select the variables that starts with `yr` or ends with `qual`. 




