## Reorder Observations {#reorder-observations}

Now, let's look at the third task: find the 10 countries with the highest life expectancy in the year 2008. To order observations, you can use the function `arrange()` in the **dplyr** package. 

First, let's create a new dataset called `gm_2008` that only contains the observations in 2008 and the variables `country`, `life_expectancy`, and `GDP_per_capita`. 

```{r}
library(r02pro)
library(dplyr)
library(tibble)
gm_2008 <- gm %>% 
  filter(year == 2008) %>%
  select(country, life_expectancy, GDP_per_capita)
```


To arrange the observations in the ascending order of the life expectancy (`life_expectancy`), you just need to add `life_expectancy` as an argument of the `arrange()` function. To arrange the descending order, you can add `desc()` around the variable. 
```{r}
gm_2008 %>% 
  arrange(life_expectancy)         #arrange in the ascending order of life_expectancy
gm_2008 %>%
  arrange(desc(life_expectancy))   #arrange in the descending order of life_expectancy
```


You may observe from the results that there are several countries with the same `life_expectancy` value, leading to a tie. To break the tie, you can supply additional variables in the `arrange()` function, which will arrange the observations within the tie according to the additional variables in the order they are supplied.

```{r}
gm_2008 %>% 
  arrange(life_expectancy, 
          GDP_per_capita) 
```

Here, the observations are arranged in the ascending order of `life_expectancy`, and the ties are broken in the ascending order of `GDP_per_capita`. Note that the observations that has an `NA` value in the specified variable will always be arranged at the end of the output. 

If you want to break the tie in the descending order of `GDP_per_capita`, you can use `desc()` around the variable. 

```{r}
gm_2008 %>% 
  arrange(life_expectancy, 
          desc(GDP_per_capita)) 
```

Now, we are ready to present the 10 countries with the highest life expectancy in 2008. 

```{r}
gm_2008 %>% 
  arrange(desc(life_expectancy)) %>% 
  head(10)
```

Here, the `head(10)` function is used to get the first 10 observations in the dataset.

### Exercises

Using the `ahp` dataset,
1. Find all houses built in `2008` with `house_style` as `2Story`, then arrange the the observations in the ascending order of remodel year, and break the ties in the descending order of `sale_price`.

2. Find all houses sold in `2009` with `house_style` as `1Story`, and arrange the observations in the descending order of `sale_price`.

