## Separate and Combine Columns via `separate()` and `unite()`  {#separate-unite-columns}

Real-world data is often messy. Sometimes, valuable information is crammed into a single column, making it difficult to analyze.  Other times, separate columns might need to be combined for specific tasks. The `separate()` and `unite()` functions in **tidyr** provide powerful tools to reshape your data and make it tidy for analysis. Imagine a dataset with a column like 'City, State'.  `separate()` allows you to split this into separate `'City'` and `'State'` columns. Conversely, `unite()` can combine `'First Name'` and `'Last Name'` columns into a single `'Full Name'` column.

To introduce the separate and and unite operation, let's again consider the following tidy data set. 
```{r}
library(r02pro)
library(tidyverse)
gm_tidy <- gm %>% 
  filter(country %in% c("United States", "China", "Russia")) %>%
  select(country, year, life_expectancy) %>%
  filter(year >= 2004 & year <= 2006)
gm_tidy
```


### Combine Columns via `unite()`

Sometimes, we may want to **combine multiple columns** into a single one. To do this, you can use the `unite()` function in the **tidyr** package. 

In the `unite()` function, you need to specify the name of the to-be created column (`col`), the columns to be combined, and the separator between the columns (`sep`). 
```{r, results=TRUE}
gm_year <- gm_tidy %>% 
  unite(col = "country_year", country, year, sep = "_")
gm_year
```

Now, the `gm_year` tibble contains a new column `country_year`, which is the combination of the `country` and `year` columns. The separator between the two columns is an underscore `_`.

### Separate One Column into Multiple Columns Using `separate()`


We know the `unite()` function combine multiple columns into a single column. Sometimes, we need to do the opposite operation, namely separate one column into multiple ones.

For example, `gm_year` is not a tidy dataset since the `country_year` column contains two variables. 

To make it tidy, you can use the `separate()` function. In the `separate()` function, you need to specify

- `col`: the name of the column to be separated
- `into`: the desired names of the to-be created columns after separation
- `sep`: the separator during the separation process


For the `country_year` column in the `gm_year` tibble, we can separate it into two columns: `country` and `year`. 

```{r}
gm_year %>% 
  separate(col = "country_year", 
           into = c("country", "year"), 
           sep = "_")
```

From the result, you can see that the original **tidy** tibble is recovered via the `separate()` function. After a careful look, you may notice that the column types of `country` and `year` are both `character`, which may not be what we want. To convert the types automatically, you can add the `convert = TRUE` argument in the `separate()` function.

```{r}
gm_year %>% 
  separate(col = "country_year", 
           into = c("country", "year"), 
           sep = "_", 
           convert = TRUE)
```

Now, the `country` and `year` columns are converted to the appropriate types.


In addition to separating via the separator, you can also separate by the position. For example, to separate the year into century (first two digits) and 2-digit year (last 2 digits), you can separate the `year` after position 2. 

```{r}
gm_tidy %>% 
  separate(col = "year", 
           into = c("y12", "y34"), 
           sep = 2, 
           convert = TRUE)
```

From the result, you can see that the `year` column is separated into two columns: `y12` and `y34`. The `y12` column contains the first two digits of the year, while the `y34` column contains the last two digits of the year.


You can also separate at multiple locations. For example, to separate the year into four columns, representing four digits, you can use multiple  positions to separate.

```{r}
gm_digits <- gm_tidy %>% 
  separate(col = "year", 
           into = c("y1", "y2", "y3", "y4"), 
           sep = c(1, 2, 3), 
           convert = TRUE)
gm_digits
```


Using the `gm_digits`, you can also combine the four columns into a single column named `year`. 

```{r}
gm_digits %>% unite(col = "year", y1:y4, sep = "")
```

Note that the `sep` argument in the `unite()` function is an empty string `""`. This means that there is no separator between the columns.

### Exercises

1. You have a tibble containing data on disease outbreaks, but the location column combines the country and city.
```{r}
outbreaks <- tibble(
  outbreak_id = c(1, 2, 3),
  disease = c("Measles", "Influenza", "Cholera"),
  location = c("USA_New York City", "Brazil_Sao Paulo", "India_Mumbai")
)
outbreaks
```

Separate the `location` column into `country` and `city` columns.

2.  You have a tibble with separate columns for patient medical record numbers and visit dates.
```{r}
patients <- tibble(
  patient_id = c(1, 2, 3),
  record_number = c("A123", "B456", "C789"),
  visit_date = c("2024-10-15", "2024-09-22", "2024-10-16")
)
patients
```

Combine the `record_number` and `visit_date` columns into a single column named `record_date` with the format "record_number - visit_date".

3. You have a tibble with a vaccine_date column in the format "YYYY-MM-DD".

```{r}
vaccinations <- tibble(
  patient_id = c(1, 2, 3),
  vaccine_name = c("MMR", "Flu", "Tdap"),
  vaccine_date = c("2023-08-10", "2024-10-05", "2022-06-15")
)
vaccinations
```

- Split the `vaccine_date` column into three columns: year, month, and day.
- Create a new column called `month_year` in the format "MM/YYYY".

4. You have a tibble with responses to a health survey, but some responses in the symptoms column are separated by commas, while others are separated by semicolons.

```{r}
survey_data <- tibble(
  respondent_id = c(1, 2, 3),
  symptoms = c("Headache, Fatigue", "Nausea; Vomiting", "Cough")
)
survey_data
```

Split the symptoms column into `symptom1` and `symptom2` columns, handling the different separators correctly.
