## Convert Between Names and Values  {#tidy-name-value}


First, let's prepare a subset of the `gm` data in the **r02pro** package. 

```{r}
library(r02pro)
library(tidyverse)
gm_tidy <- gm %>% 
  filter(country %in% c("United States", "China", "Russia")) %>%
  select(country, year, life_expectancy) %>%
  filter(year >= 2004 & year <= 2006)
gm_tidy
```

From the code, you can see that `gm_tidy` is a tibble that contains the life expectancy of the United States, China, and Russia from 2004 to 2006. This is a **tidy** format.

### Convert Values into Column Names {#tidy-wide}

In `gm_tidy`, each row contains the year when the life expectancy measurement was taken. Suppose we want to convert the year value into column names. You can use the `pivot_wider()` function.

In `pivot_wider()`, you need to specify two arguments: 

- `names_from` denotes which column in the original tibble contains the values of the new column names, 
- `values_from` denotes which column in the original tibble contains the values for each cell in the new tibble. 

```{r}
gm_wide <- gm_tidy %>% 
  pivot_wider(names_from = year, 
              values_from = life_expectancy)
gm_wide
```

From the resulting tibble `gm_wide`, you can see that the `year` variable is converted into column names. The `life_expectancy` variable contains the corresponding values. The `gm_wide` is now in a wide format. During the process, the `year` information is lost, which is not desirable. Fortunately, you can add a prefix "year" to the column names via an argument `names_prefix` in the `pivot_wider()` function. 

```{r}
gm_wide_year <- gm_tidy %>% 
  pivot_wider(names_from = year, 
              names_prefix = "year", 
              values_from = life_expectancy)
gm_wide_year
```

In the resulting tibble `gm_wide_year`, you can see that the column names are prefixed with "year". The `year` information is now reflected in the names of the tibble.

The reason why we call the function `pivot_wider()` is that it will create a **wider** dataset than the original one, containing more columns.

### Convert Column Names into Values {#tidy-long}

Apparently, both `gm_wide` and `gm_wide_year` are not tidy. Now, let's make them tidy using the `pivot_longer()` function. 
```{r}
gm_wide %>% 
  pivot_longer(cols = -1, 
               names_to = "year", 
               values_to = "life_expectancy")
```


In the `pivot_longer()` function, 

- `cols` specifies the column names that you want to convert from, which accept the same format as that in `dplyr::select()` introduced in Section \@ref(select-variables). 
- `names_to` specifies the variable name you want to use for the column names. 
- `values_to` specifies the variable name for holding the values in the selected columns.

In the resulting tibble, `col = -1` means that all columns except the first column are selected. The column names `2004`, `2005`, and `2006` are converted into values of the variable `year`. The `life_expectancy` variable contains the corresponding values. We have recovered the tidy `gm_tidy` through the tidy process.

The reason why we call the function `pivot_longer()` is that it will create a **longer** dataset than the original one, containing more rows.

Now, let's tidy `gm_wide_year` using the `pivot_longer()` function. 

```{r}
gm_wide_year %>% 
  pivot_longer(cols = -1, 
               names_to = "year", 
               names_prefix = "year", 
               values_to = "life_expectancy")
```

Note that we have an extra argument `names_prefix` in the `pivot_longer()` function. This argument is used to remove the prefix "year" from the column names.

Comparing the `pivot_wider()` and `pivot_longer()` functions, you can see that they are complementary. The `pivot_wider()` function converts values into column names, while the `pivot_longer()` function converts column names into values.

### Multiple Variables {#tidy-multiple}

In the `gm` dataset, there are multiple variables that you may want to convert into column names. You can specify multiple variables in the `names_from` argument in the `pivot_wider()` function. 

```{r}
gm_tidy <- gm %>% 
  filter(country %in% c("United States", "China", "Russia")) %>%
  select(country, year, life_expectancy, GDP_per_capita) %>%
  filter(year >= 2004 & year <= 2006)
gm_tidy
gm_wide <- gm_tidy %>% 
  pivot_wider(names_from = c(country, year), 
              values_from = c(life_expectancy, GDP_per_capita),
              names_sep = "-")
gm_wide
```

In the `pivot_wider()` function, you can specify multiple variables in the `names_from` and `values_from` arguments. The resulting tibble `gm_wide` contains the life expectancy and GDP per capita of the United States, China, and Russia from 2004 to 2006. The column names are in the format of `country_year_variable`. Note that we have an extra argument `names_sep` in the `pivot_wider()` function. This argument is used to separate the column names with a hyphen `"-"`. We didn't use the separator `"_"` because it is already used in the column names of the original tibble.

To make the tibble tidy, you can use the `pivot_longer()` function. 

```{r}
gm_wide %>% 
  pivot_longer(cols = everything(), 
               names_to = c(".value", "country", "year"), 
               names_sep = "-")
```

- Here, we use the `everything()` function to select all columns. 
- The `names_to` argument specifies the variable names for the column names. The `".value"` is a special value that tells the function to keep the values in the selected columns. The `country` and `year` variables are used to store the country and year values in the column names.
- The `names_sep` argument specifies the separator used in the column names. 

Now, let's see another example.
```{r}
gm_gender <- gm %>% 
  filter(country %in% c("United States", "China", "Russia")) %>%
  filter(year == 2004) %>%
  select(country, ends_with("male"))
gm_gender
```

Here, we have selected the columns that end with `male`. 
One tricky part is that some columns contain more than one underscores, making the pattern a bit more complicated.

We want to convert the column names into values. 
```{r}
gm_gender %>% 
  pivot_longer(cols = -1,
               names_to = c(".value", "gender"),
               names_pattern = "(.+)_(?!.*_)(.+)")
```


In the `pivot_longer()` function, the `names_to` argument specifies the variable names for the column names. Here, the `".value"` is a special value that tells the function to keep the values in the selected columns. The `gender` variable is used to store the values of the column names. The `names_pattern` argument specifies the pattern of the column names. The `(.+)_(?!.*_)(.+)` regular expression uses a negative lookahead assertion to identify the last underscore. In particular,

- `(.+)`: Captures everything before the last underscore.
- `_(?!.*_)`: Matches an underscore only if it's not followed by another underscore.
- `(.+)`: Captures everything after the last underscore.


### Exercises

1. You are given the following dataset. 
```{r}
df <- tibble(
  country = c("A", "A", "B", "B"),
  year = c(2000, 2001, 2000, 2001),
  life_expectancy = c(70, 71, 75, 76),
  infant_mortality = c(5, 4.5, 3, 2.8)
)
```

Create a wide-format dataset where:

- Each row represents a country.
- Columns represent combinations of year and a demographic variable (e.g., '2000_life_expectancy', '2000_infant_mortality', '2001_life_expectancy', etc.).

2. Use the wide-format dataset created in Exercise 1, transform the data back into a tidy format.

3. You are given the following dataset. 
```{r}
df <- tibble(
  country = c("A", "B", "C"),
  life_expectancy_male = c(72, 78, 75),
  life_expectancy_female = c(78, 83, 80)
)
```
Tidy the dataset so that it contains the following columns: `country`, `gender`, and `life_expectancy`.

4. You are given the following dataset. 
```{r}
df <- tibble(
  group = c("A", "B"),
  q1_pre = c(10, 12),
  q1_post = c(15, 18),
  q2_pre = c(7, 9),
  q2_post = c(11, 13)
)
```

Tidy the dataset so that it contains the following columns: `group`, `question`, `time`, and `score`, where `question` is either `q1` or `q2`, `time` is either `pre` or `post`, and `score` is the corresponding value.
