## Matching Pattern with Regular Expressions {#match-pattern}

In many applications, we may want to find out which strings match a certain pattern.  To illustrate this, we will be using the variable names from the `sahp` data set from **r02pro** package. 

```{r}
library(r02pro)
library(stringr)
sahp_names <- colnames(sahp)
```

The **stringr** package provides a very useful function called `str_view()` to highlight the elements that match the given pattern. To create a logical vector that reflect this match, you can use the `str_detect()` function.

### Basic Matches

Let's describe a few commonly used patterns.

***a. contain a given string***

To find all strings that contain `"room"`, you can run the following code.
```{r, results = TRUE}
str_view(sahp_names, "room")
str_detect(sahp_names, "room")
```

You can see that the `str_detect()` function returns a logical vector that reflects whether this is a match. In the rest of this section, we will be focusing on demonstration with `str_view()`.

From the result, we can see that `"bedroom"` and `"bathroom"` are highlighted, since they both contain the string `"room"`. 

***b. including `.` in the pattern***

In addition to specifying the exact string for matching, you can also use the wild symbol `.` to represent **any character**. 
```{r, results=TRUE}
str_view(sahp_names, ".h")
str_view(sahp_names, ".s.")
```

The first pattern `".h"` represents any character plus the letter h. Note that `"house_style"` is not a match since it doesn't have any character before the letter `"h"`.  

The second patter `".s."` represents a length-3 substring, with any two characters around `"h"`. Again, here `"sale_price"` is not a match since `"s"` is a first character. 

***c. including anchors in the pattern***

Sometimes, we may want to match the **start** of the string (`"^"`), or the **end** of the string (`"$"`).
```{r, results = TRUE}
str_view(sahp_names, "^l")
```

The highlighted strings `"liv_area"` and `"lot_area"` both start with letter `"l"`. Although `"oa_qual"` and `"house_style"` contain `"l"`, they are not selected since `"l"` is not the start of the strings. Now, let's find the strings the end with  `"l"`.

```{r, results=TRUE}
str_view(sahp_names, "l$")
```

We now see `"oa_qual"`, `"kit_qual"`, and `"heat_qual"` all selected. 


### Character Classes Matches

***a. Single character match***

Sometimes, we may want to match the characters with a specific class or a group of values instead of exact values. Here is a list of the commonly used classes.

- `\d`: matches any digit. In other words, numbers from 0 to 9.
- `\s`: matches any whitespace (e.g. space, tab, newline). For example, ` `, `\t`, `\n`.
- `[xyz]`: matches x, y, or z.
- `[^xyz]`: matches anything except x, y, or z. (The opposite of `[xyz]`)
- `[a-z]` or `[:lower:]`: matches every character between a and z.
- `[A-Z]` or `[:upper:]`: matches every character between A and Z.

Note that all of these are detecting a single character from the candidate ones.

Let's see some examples.
```{r, results=TRUE}
my_char <- c("abc","a1","2b","33c","d 2", "d f  3")
str_view(my_char, "\\d")
```

First of all, we need to have `"\\"` since `"\"` is a special character that needs the **espace** character `"\"`. Here, as long as the string contains at least one digits, it will be matched.

```{r, results=TRUE}
str_view(my_char, "\\s\\d")
```

Here, the pattern `"\s\d"` represents a pattern of a whitespace followed by a digit. Here, only `"d 2"` and `"d f  3"` match this pattern. 


```{r, results=TRUE}
str_view(my_char, "[ac3]")
```

Here, `"[ac3]"` indicates that as long as the string contains `"a"`, `"c"`, or `"3"`, it will be matched. 

```{r, results=TRUE}
str_view(my_char, "[^abc3]")
```

Here, `"[^abc3]"` indicates that as long as the string contains any character that is not one of `"a"`, `"b"`, `"c"`, and `"3"`, it will be matched. 

***b. Repeated characters match***
Sometimes, you may want to match a pattern multiple times consecutively. While you can do it manually by repeating the pattern, there is an easier way to do this by using the `{}` with the number inside. Let's create a string and see some examples.
```{r}
my_str_rep <- c("bbbb", "1234a",
                "bbb", "abcdef",
                "bb ee", "123456789")
```


`b{3}` will match 3 b's. 
```{r}
str_view(my_str_rep, "b{3}")
```

To match a 9-digit number, you can use `[1-9]{9}`.
```{r,results=TRUE}
str_view(my_str_rep, "[1-9]{9}")
```


`[a-z]{4}` matches all four-letter word.
```{r, results = TRUE}
str_view(my_str_rep, "[a-z]{4}")
```
`[a-z1-9]{4,}` will match all strings which has at least 4 lower case letters or numbers.  

```{r, results = TRUE}
str_view(my_str_rep, "[a-z1-9]{4,}")
```





