## Other Distributions {#other-distributions}


In Section \@ref(normal-distribution), we gave a detailed introduction to the four functions for a normal distribution, which is a popular *continuous* distribution. In particular, we now know that `dnorm()` produces the pdf of a normal distribution. In the case of *discrete* distributions, however, we would have **probability mass function (pmf)** instead of the pdf. Let's use the **binomial** distribution as a representative example of discrete distributions with the four functions as below.

```{r echo = FALSE, results = TRUE}
Code  <- c("`dbinom(3, size, prob)`","`pbinom(3, size, prob)`","`qbinom(3, size, prob)`","`rbinom(3, size, prob)`")
Name <- c("probability mass function", "cumulative distribution function", "quantile function", "random number generator")
d <- data.frame(Code, Name)
knitr::kable(d)
```


Now, let's look at a few other commonly used  distributions. For simplicity, let's just use the random number generator for each distribution in the following table. 

```{r echo = FALSE, results = TRUE}
Name <- c("exponential", "uniform", "t", "F", "beta","gamma", "poisson","binomial","bernoulli")
Code  <- c("`rexp(3, rate = 0.5)`","`runif(3, min = 1, max = 2)`","`rt(3, df = 4)`","`rf(3, df1 = 3, df2 = 6)`", "`rbeta(3, shape1 = 2, shape2 = 3)`", "`rgamma(3, shape = 2, rate = 3)`", "`rpois(3, lambda = 5)`", "`rbinom(3, size = 3, prob = 0.3)`", "`rbinom(3, size = 1, prob = 0.5)`")
para_1 <- c("rate", "min", "df", "df1", "shape1", "shape", "lambda","size","size")
para_2 <- c("-", "max", "-", "df2", "shape2", "rate", "-", "prob", "prob")
d <- data.frame(Name, Code, para_1, para_2)
knitr::kable(d)
```

As we can see from this table, all random number generator functions are formed by the letter `r` followed by the name of the distribution we would like to generate from. For the other three functions, we just need to change the initial letter `r`: 

- to `d` for pdf (continuous distribution) or pmf (discrete distribution), 
- to `p` for cdf,
- to `q` for quantile function.

Let's do some statistical exercises with those distributions.

### Exercise

