## Concatenate and Subset Strings {#combine-subset-strings}

### Concatenate Strings

First, let's introduce how to **concatenate multiple strings** into a single string. To do this, we will use the `str_c()` function in the **stringr** package.

```{r, results=TRUE}
library(stringr)
str_c("I", "am", "learning", "R!")    #equivalent to `paste0()`
```

Similar to the `paste()` function, you can specify a separator between the input strings. 

```{r, results=TRUE}
str_c("I", "am", "learning", "R!", sep = " ")
str_c("apple", "banana", "cherry", sep = "&")
```

The `str_c()` can also accept vectors of length greater than one as its arguments, where it will conduct the concatenating operation  elementwisely.
 
```{r, results=TRUE}
str_c("There are", 3:5, c("apples", "bananas", "cherries"), "in the kitchen.", sep = " ") 
```

As you can see from the result, the first and third string are recycled to match the length of the second string. 
 

Next, let's introduce how to concatenate the strings inside a chracter vector. First, let's try to apply the `str_c()` function on a character vector.
 
```{r, results=TRUE}
fruits <- c("apples", "bananas", "cherries")
str_c(fruits)
```
 

Apparently, we didn't get the expected concatenated string. To do so, we need to add an argument `collapse` as the separator.

```{r, results=TRUE}
str_c(fruits, collapse = "|")
str_c(fruits, collapse = "-")
```

When we have multiple character vectors as the arguments in the `str_c()` function with the `collapse` argument, the character vectors will first be concatenated respectively, and then the concatenated strings will be further concatenated with the separator as in the `collapse` argument.

```{r, results=TRUE}
str_c(fruits, fruits, sep = '_', collapse = "|")
```

Here, the `str_c()` function will first concatenate the `fruits` vector with itself with the separator `_`, and then concatenate the two concatenated strings with the separator `|`. 

### Create and Modify Substrings

It is often of interest to create **substrings** from an existing string. A **substring** is a string formed by a consecutive sequence of characters within the existing string.



To get a substring, you can use the `substr()` function. The `start` and `end` arguments represent the starting and ending position of the desired substrings. 

```{r, results = TRUE}
str_sub(fruits, start = 1, end = 3)
```

Here, you will get a substring from position 1 to position 3 for each element of the original character vector. In addition to using the positive integer to represent the position, you can also use negative integer to represent the relative position from the last symbol.

```{r, results=TRUE}
str_sub(fruits, start = 1, end = -2)
```

Here, for each string, we extract the substring from position 1 to the second to last position. In addition to using one integer for start and end, you can also supply a vector of the same length as the input vector to extract multiple substrings at once.

```{r, results=TRUE}
str_sub(fruits, start = 1:3, end = (-1):(-3))
```

Let's look at the result together.

From the results, we can see that the substrings are extracted as follows:

- The first element is the first to the last position of `"apples"`.
- The second element is the second to the second to last position of `"bananas"`.
- The third element is the third to the third to last position of `"cherries"`.

Once we know how to create the substring, it is straightforward to modify the substing in place of the original string. Let's try to update the substrings.
```{r}
fruits_new <- fruits
str_sub(fruits_new, start = 1:3, end = (-1):(-3)) <- 
        c("ns_1", "ns_2", "ns_3")
fruits_new
fruits
```

From the results, we can see the corresponding substrings are replaced by `"ns_1", "ns_2", "ns_3"`, respectively.


### Exercises

1. (String Concatenation) Create a character vector `fruits` with the elements `"apple"`, `"banana"`, and `"cherry"`. Use the `str_c()` function to concatenate the elements in the `fruits` vector with the separator `" & "`.

2. (Create Substrings) Create a character vector `colors` with the elements `"red"`, `"green"`, `"blue"`, `"yellow"`, and `"purple"`. Use the `str_sub()` function to extract the substrings from the second to the fourth position of each element in the `colors` vector.

3. (Update Substring) Create a character vector `animals` with the elements `"cat"`, `"dog"`, `"elephant"`, and `"giraffe"`. Use the `str_sub()` function to update the substrings from the second to the fourth position of each element in the `animals` vector to `"at"`, `"og"`, `"leph"`, and `"iraf"`, respectively. Verify the contents.


