## Normal Distribution  {#normal-distribution}

First, let's review the definition of **normal distribution**, which is also called **Gaussian distribution**. If $X\sim N(\mu, \sigma^2)$, we say $X$ is a random variable following a normal distribution with mean $\mu$ and variance $\sigma^2$. 

In the following table, we list the four useful functions for normal distribution, and they will be introduced in the subsequent four parts, respectively.

```{r echo = FALSE, results = TRUE}
Code  <- c("`dnorm(x, mean, sd)`","`pnorm(q, mean, sd)`","`qnorm(p, mean, sd)`","`rnorm(n, mean, sd)`")
Name <- c("probability density function", "cumulative distribution function", "quantile function", "random number generator")
Section <- c("\\@ref(pdf)","\\@ref(cdf)","\\@ref(qf)","\\@ref(rng)")
d <- data.frame(Code, Name, Section)
knitr::kable(d)
```




### Probability Density Function (pdf) {#pdf}

To characterize the distribution of a continuous random variable, you can use the **probability density function (pdf)** . When $X\sim N(\mu,\sigma^2)$, its pdf is
$$f(x) = \frac{1}{\sqrt{2\pi \sigma}}\exp\left[-\frac{(x-\mu)^2}{2\sigma^2}\right].$$

In R, you can use `dnorm(x, mean, sd)` to calculate the pdf of normal distribution. 

- The argument `x` represent the location(s) at which to compute the pdf. 

- The arguments `mean` and `sd` represent the mean and standard deviation of the normal distribution, respectively. 


For example, `dnorm(0, mean = 1, sd = 2)` computes the pdf at location 0 of $N(1, 4)$, normal distribution with mean 1 and variance 4. 

::: {.infobox .caution data-latex="{caution}"}
Note that the argument `sd` is the standard deviation, which is the square root of the variance. 
::: 

In particular, `dnorm()` without specifying the `mean` and `sd` arguments will compute the pdf of $N(0,1)$, which is the standard normal distribution. Let's see  examples of computing the pdf at one location for three different normal distributions.

```{r, results = TRUE}
dnorm(0, mean = 1, sd = 2)
dnorm(1, mean = -1, sd = 0.5)
dnorm(0) #standard normal
```

In addition to computing the pdf at one location for a single normal distribution, `dnorm` also accepts vectors with more than one elements in all three arguments. For example, you can use the following code to compute the three pdf values in the previous code block.
```{r, results = TRUE}
dnorm(c(0,1,0), mean = c(1, -1, 0), sd= c(2, 0.5, 1))
```

If you want to compute the pdf at the same location 0 for distributions $N(1,4)$, $N(-1, 0.25)$, and $N(0, 1)$, you can use the following code. 
```{r, results = TRUE}
dnorm(0, mean = c(1, -1, 0), sd= c(2, 0.5, 1))
```

If you want to compute the pdf at three different locations (-3, 2, and 5) for distribution $N(3, 4)$, you can use the following code. 

```{r, results = TRUE}
dnorm(c(-3, 2, 5), mean = 3, sd = 2)
```


To get a better understanding on the shape of the normal pdf, let's visualize the pdf of $N(0,1)$. You first need to create a equal-spaced vector `x` from -5 to 5 with increment 0.1. Then, you can compute the pdf value for each element of `x` using `dnorm()`. Finally, you can visualize the pdf using `geom_line()`.

```{r, results = TRUE}
library(ggplot2)
x <- seq(from = -5, to = 5, by = 0.05)
norm_dat <- data.frame(x = x, pdf = dnorm(x))
ggplot(norm_dat) + geom_line(aes(x = x, y = pdf))
```

Next, you can take a step further to visualize three different normal distributions in the same plot, $N(0,1)$, $N(1,4)$, and $N(-1, 0.25)$. You can use the same vector `x` and compute the three pdfs on each element of `x`. `geom_line()` is still used with the variable `dist` mapped to the `color` aesthetic. 

```{r, results = TRUE}
x <- seq(from = -5, to = 5, by = 0.05)
norm_dat_1 <- data.frame(dist = "N(0,1)", x = x, pdf = dnorm(x))
norm_dat_2 <- data.frame(dist = "N(1,4)", x = x, pdf = dnorm(x, mean = 1, sd = 2))
norm_dat_3 <- data.frame(dist = "N(-1, 0.25)", x = x, pdf = dnorm(x, mean = -1, sd = 0.5))
norm_dat <- rbind(norm_dat_1, norm_dat_2, norm_dat_3)
ggplot(norm_dat) + geom_line(aes(x = x, y = pdf, color = dist))
```

### Cumulative Distribution Function (cdf) {#cdf}

In addition to pdf, you can compute the **cumulative distribution function (cdf)** of the normal distribution using the function `pnorm(q, mean, sd)`. Generally speaking, the cdf of a random variable $X$ is defined as 
$$F(x) = P(X\leq x).$$ Similar to `dnorm()`,  `pnorm()` also has two optional arguments, `mean` and `sd`, which represent the mean and standard deviation of the normal distribution, respectively. If you don't specify these two arguments, `pnorm()` will compute the cdf of $N(0,1)$. 
```{r, results = TRUE}
pnorm(0, mean = 1, sd = 2)
pnorm(0) # cdf at 0 of standard normal
```

You can also use `pnorm()` to visualize the cdf of the standard normal distribution.

```{r, results = TRUE}
q <- seq(from = -5, to = 5, by = 0.1)
norm_dat <- data.frame(q = q, cdf = pnorm(q))
ggplot(norm_dat) + geom_line(aes(x = q, y = cdf))
```

### Quantile Function  {#qf}

The third useful function related to distributions is the **quantile function**. You can compute the quantile of the normal distribution using `qnorm(p, mean, sd)`. The quantile function is the inverse function of the cdf. In particular, the $p$ quantile returns the value $x$ such that
$$F(x) = P(X\leq x) =  p$$ 

Let's verify `qnorm()` is indeed the inverse function of `pnorm()` using the following example. 

```{r, results = TRUE}
pnorm(qnorm(c(0.5,0.7)))
```


When $p=0.5$, `qnorm()` gives us the median of the normal distribution. 
Let's see a few examples for computing the quantiles. 
```{r, results = TRUE}
qnorm(0.5, mean = 1, sd = 2)
qnorm(0.5)
```



You can also visualize the shape of the quantile function.

```{r, results = TRUE}
p <- seq(from =  0.01, to = 0.99, by = 0.01)
norm_dat <- data.frame(p = p, quantile = qnorm(p))
ggplot(norm_dat) + geom_line(aes(x = p, y = quantile))
```



### Random Number Generator  {#rng}

Lastly, to generate (pick up) random numbers from normal distributions, you can use the function `rnorm(n, mean, sd)` , with the argument `n` represents the number of random numbers to generate, the arguments `mean` and `sd` are the mean and standard deviation of the normal distribution you would like to generate from, respectively. Again, if you only supply the argument `n`, you will be generating random numbers from $N(0,1)$. 

```{r, results = TRUE}
rnorm(3, mean = 0, sd = 1) #generate 3 random numbers from N(0, 1)
rnorm(3) #generate another 3 random numbers from N(0,1)
```

Since you are generating random numbers, the results may be different each time. In many applications, however, you may want to make the results reproducible. To do this, you can set random seed using the function `set.seed()` before generating the random numbers. Let's see the following example.

```{r, results = TRUE}
set.seed(0)
rnorm(3) 
```

Now, let's run it one more time. 

```{r, results = TRUE}
set.seed(0)
rnorm(3) 
```

You can see that the exact 3 numbers are reproduced since you are using the same random seed 0. You can run these two lines of code on any machine and will get the exact same three random numbers. 

::: {.infobox .caution data-latex="{caution}"}
Note that the code that involves randomness needs to be identical to reproduce the results. If you change the arguments in `rnorm()`, you will get totally different results. See the following example. 
::: 

```{r, results = TRUE}
set.seed(0)
rnorm(1)
rnorm(3) 
```

By setting a different random seed, you will see different results as the following example.

```{r, results = TRUE}
set.seed(1)
rnorm(3) 
```


Lastly, let's do a simple statistical exercise by checking the closeness of the *sample mean* and *sample standard deviation* to their population counterparts. 

```{r, results = TRUE}
x <- rnorm(1e6, mean = 1, sd = 2)
mean(x) #sample mean
var(x)  #sample covariance
sd(x)   #sample standard deviation
```

### Exercise

