## Matching Pattern with Regular Expressions {#match-pattern}

Regular expressions provide a powerful and flexible way to find and manipulate text based on patterns of characters. They are essential tools for various tasks in data analysis and programming, including:

-   **Data cleaning**: Standardizing formats of names, addresses, or phone numbers.
-   **Validation**: Ensuring user input adheres to expected patterns (e.g., email addresses).
-   **Text extraction**: Isolating keywords, dates, or other crucial information from documents.

This section introduces you to the fundamentals of regular expressions in R, using the **stringr** package.

```{r}
library(stringr)
```

The **stringr** package provides a very useful function called `str_view()` to highlight the elements that match the given pattern. To create a logical vector that reflect this match, you can use the `str_detect()` function.

### Basic Pattern Matching

We'll utilize a sample character vector to illustrate these concepts.

```{r}
sample_names <- c("apple_pie", "banana_bread", "cherry_pie", "chocolate_cake", "apple_crumble")
```

Let's describe a few commonly used patterns.

***a. matching literal strings***

The simplest form of pattern matching involves searching for a specific sequence of characters. To find all variable names containing the exact string `"pie"`, you can run the following.

```{r, results = TRUE}
str_view(sample_names, "pie")
```

Here, the `str_view()` function highlights the variable names that contain the string `"pie"`.

```{r}
str_detect(sample_names, "pie")
```

The `str_detect()` function returns a logical vector that reflects whether this is a match. In the rest of this section, we will be focusing on demonstration with `str_view()`.

***b. using wildcards***

The period `(.)` acts as a wildcard, matching **any** single character. This allows for more flexible pattern definitions. For example, to find all variable names that contain the letter `"c"` followed by any character, you can use the pattern `".c"`.

```{r, results=TRUE}
str_view(sample_names, ".c")
str_view(sample_names, "._pi.")
```

The first pattern `".p"` represents any character plus the letter c. Note that `"cherry_pie"` is not a match since it doesn't have any character before the letter `"c"`.

The second pattern `"._pi."` represents any character plus the string `"_pi"` plus any character.

***c. including anchors***

Anchors allow you to specify the position of a match within a string: - `^` matches the start of a string. - `$` matches the end of a string.

For example, to find all variable names that start with the letter `"c"`, you can use the pattern `"^c"`.

```{r}
str_view(sample_names, "^c")
```

Now, let's find the strings the end with `"pie"`.

```{r}
str_view(sample_names, "pie$")
```

### Character Classes

***a. matching single characters***

Sometimes, we may want to match the characters with a specific class or a group of values instead of exact values. Here is a table of the commonly used classes.

```{r, echo=FALSE}
library(kableExtra)

df <- data.frame(
  Class = c("\\\\d", "\\\\s", "[xyz]", "[^xyz]", "[a-z] or [:lower:]", "[A-Z] or [:upper:]"),
  Description = c(
    "Matches any digit (0-9)", 
    "Matches any whitespace character (space, tab, newline)",
    "Matches any one of the characters within the brackets (x, y, or z)",
    "Matches any character except those within the brackets",
    "Matches any lowercase letter", 
    "Matches any uppercase letter"
  )
)

kable(df, "html", escape = FALSE) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

Let's see some examples.

```{r}
my_char <- c("abc","a1","2b","33c","d 2", "d f  3")
str_view(my_char, "\\d")       # Matches strings containing a digit
str_view(my_char, "\\s\\d")    # Matches a whitespace followed by a digit
str_view(my_char, "[ac3]")     # Matches strings containing 'a', 'c', or '3'
str_view(my_char, "[^abc3]")  # Matches strings containing characters other than 'a', 'b', 'c', or '3'
```

Here, `\\d` indicates that as long as the string contains a digit, it will be matched. As a result, `"a1"`, `"2b"`, and `"33c"` are matched. Note that you need to escape the backslash `\` by using two backslashes `\\` within R strings.

***b. matching repeated characters***

To match a character or group of characters multiple times in a row, you can use quantifiers:

```{r, echo=FALSE}
library(knitr)
library(kableExtra)

df <- data.frame(
  Symbol = c("{n}", "{n,}", "{n,m}", "?", "\\+", "\\*"),
  Description = c(
    "Matches exactly n occurrences of the preceding element.",
    "Matches n or more occurrences.",
    "Matches at least n and at most m occurrences.",
    "Matches zero or one occurrence (equivalent to {0,1}).",
    "Matches one or more occurrences (equivalent to {1,}).",
    "Matches zero or more occurrences (equivalent to {0,})."
  )
)

kableExtra::kable(df)
```

Let's see some examples.

```{r}
my_str_rep <- c("bbbb", "1234a", "bbb", "abcdef", "bb ee", "123456789")

str_view(my_str_rep, "b{3}")       # Matches consecutive three 'b's
```

Here, `"b{3}"` indicates that the string should contain exactly three `"b"`s. As a result, `"bbbb"` and `"bbb"` are matched.

```{r}
str_view(my_str_rep, "[1-9]{9}")    # Matches a 9-digit number
```

Here, `"[1-9]{9}"` indicates that the string should contain exactly nine digits. As a result, `"123456789"` is matched.

```{r}
str_view(my_str_rep, "[a-z]{4}")    # Matches four consecutive lowercase letters
```

Here, `"[a-z]{4}"` indicates that the string should match exactly four consecutive lowercase letters. As a result, `"bbbb"` and `"abcdef"` are matched. Note that `"bb ee"` is not matched due to the space.

```{r}
str_view(my_str_rep, "[a-z1-9]{4,}") # Matches at least four consecutive lowercase letters or digits
```

Finally, `"[a-z1-9]{4,}"` indicates that the string should contain at least four consecutive lowercase letters or digits. As a result, `"bbbb"`, 1234a"`,`"abcdef"`, and`"123456789"\` are matched.

### Combining Patterns with Logical Operators

You can combine multiple patterns using logical operators:

```{r, echo=FALSE}
library(kableExtra)

df <- data.frame(
  Symbol = c("\\|", "&", "()"),
  Description = c(
    "Matches either the pattern on the left or the right (OR).",
    "Requires both patterns to match (AND).",
    "Groups patterns to control the scope of operators."
  )
)

kableExtra::kable(df)
```

Let's see some examples.

```{r}
my_str_comb <- c("abc", "123", "abc123", "123abc", "abc123def")
str_view(my_str_comb, "abc|123")    # Matches either 'abc' or '123'
```

Here, `"abc|123"` indicates that the string should contain either `"abc"` or `"123"`. As a result, `"abc"`, `"123"`, `"abc123"`, `"123abc"`, and `"abc123def"` are matched.

```{r}
str_view(my_str_comb, "abc&123")    # Matches 'abc' and '123'
```

Here, `"abc&123"` indicates that the string should contain a substring that matches both `"abc"` and `"123"`, which is impossible. As a result, no matches are found.

```{r}
str_view(my_str_comb, "(abc|123)abc")  
```

Here, `"(abc|123)abc"` indicates that the string should contain either `"abc"` or `"123"` followed by `"abc"`. As a result, `"abc123"` is matched.

```{r}
str_view(my_str_comb, "abc|123abc") 
```

Here, `"abc|123abc"` indicates that the string should contain either `"abc"` or `"123abc"`. As a result, `"abc"`, `"abc123"`, and `"abc123def"` are matched.

Comparing the last two examples, the parentheses `()` are used to group the patterns to control the scope of the operators.

### Conclusion

Regular expressions provide a powerful and flexible way to find and manipulate text based on patterns of characters. In this section, we introduced you to the fundamentals of regular expressions in R, using the **stringr** package. We demonstrated how to match literal strings, use wildcards, include anchors, character classes, and combine patterns with logical operators.

### Exercises

1. Matching Literal Strings:

Create a character vector called `fruits` containing the following elements: `"apple"`, `"banana"`, `"strawberry"`, `"blueberry"`, `"blackberry"`. Use `str_view()` to highlight the fruits containing the substring `"an"`.

2. Using Wildcards

Using the `fruits` vector from Exercise 1, use a wildcard to find and highlight fruits that have any character followed by `"e"`.

3. Including Anchors

Using the `fruits` vector, find and highlight fruits that ends with the string `"berry"`.



4. Character Classes

Create a new character vector called `codes` with the elements: `"A123"`, `"B456"`, `"Bb56"`, `"C789"`, `"D012"`, `"e345"`. Use a regular expression to highlight the elements that contains an uppercase letter followed by a number. 

5. Matching Repeated Characters

Create a vector named `phone_numbers` with these elements: `"555-123-4567"`, `"(555) 123 4567"`, `"5551234567"`, `"1-555-123-4567"`, `"555.123.4567"`.

- Find and highlight all phone numbers that contain exactly three digits followed by a hyphen, another three digits, a hyphen, and finally four digits (e.g., `555-123-4567`).
- Use `str_view()` to find all phone numbers that have parentheses around the first three digits (e.g., `(555) 123 4567`).

6. Email address checker

Now, you are asked to check whether a string is a valid .com email address according to the following rules

- Format: `username@domain.com`
- Username: The part before the `"@"` symbol. It is a string of length at least one that contains any combinations of letters (upper or lower cases), numbers, periods, and underscores. 
- Domain: It is a string of length at least one that contains any combination of letters and numbers. 
- `.com`: The domain should end with `.com`.

Use the `str_detect()` function with your pattern on the following strings: 
```{r}
emails <- c(
  "user.name@domain.com", 
  "user123@domain.com", 
  "user_name@domain.com", 
  "username@domaincom", 
  "username@domain.net", 
  "user@name@domain.com",
  "@icloud.com",
  "user@.com",
  "user@do_main.com"
)
```



7. Extracting Dates

Create a character vector called `dates` with the following elements: `"2021-01-01"`, `"12-31-2021"`, `"2022-28-02"`, `"2022-02-29"`, `"2022-03-01"`. Use a regular expression to highlight the dates that are in the format `YYYY-MM-DD`. (You just need to make sure the month (MM) is from 01-12, and the day (DD) is from 01-31.)







