## Multivariate Normal Distribution   {#multivariate-normal}

In this section, we introduce how to work with multivariate normal distribution in R.

First, let's review the definition of a multivariate normal distribution. Suppose a $p$-dimensional random vector ${\bf x} \sim N({\bf \mu}, {\bf \Sigma})$, where ${\bf \mu}$ is the mean vector and ${\bf \Sigma}$ is the covariance matrix. We have 
$$
f({\bf x}) = \frac{1}{(2\pi)^{p/2}\sqrt{|{\bf \Sigma}|}}\exp\left[-\frac{1}{2}({\bf x}-{\bf\mu})^T{\bf \Sigma}^{-1}({\bf x}-{\bf\mu})\right].
$$

### Generate Multivate Normal Distribution

To randomly generate ${\bf x} \sim N({\bf \mu}, {\bf \Sigma})$, you can use the `mvrnorm(n = 1, mu, Sigma)` function in the **MASS** package, which is preloaded in the base R. The function `mvnorm()` takes three arguments.

- `n`: targeted sample size
- `mu`: mean vector
- `Sigma`: covariance matrix

Let's take a look at an example where the distribution is 
$\bf\mu = (1, 2)^T$ and 
$${\bf \Sigma} = \left[\begin{matrix}
1&0.5\\
0.5&1
\end{matrix}\right]$$

```{r, results=TRUE}
library(MASS)
n <- 10
mu <- c(1, 2)
Sigma <- matrix(c(1, 0.5, 0.5, 1), 2, 2)
x <- mvrnorm(n, mu, Sigma)
x
```

`x` contains $n=10$ two-dimensional normal distributed variable, where each row represents one observation.

### Sample Mean and Sample Covariance Matrix

In practical applications, we usually have a set of observations and would like to get a sample estimate of the parameters ${\bf \mu}$ and ${\bf \Sigma}$. 

For the ${\bf \mu}$, you can directly use the sample mean as the estimate which is unbiased. 
$$\hat {\bf\mu} = [\hat\mu_j],$$
where $$\hat\mu_j = \frac{1}{n}\sum_{i=1}^nx_{ij}.$$
```{r,results=TRUE}
mu_hat <- apply(x, 2, mean)
mu_hat
```

To estimate ${\bf\Sigma}$, we will use the sample covariance matrix estimate:
$$\hat{\bf \Sigma} = [\hat\Sigma_{jk}],$$
where $$\hat\Sigma_{jk} = \frac{1}{n-1}\sum_{i=1}^n(x_{ij}-\hat\mu_j)(x_{ik}-\hat\mu_k).$$

To compute this matrix, we can use the function `cov()` on the matrix `x`.
```{r, results=TRUE}
sigma_hat <- cov(x)
sigma_hat
1/(n-1)*sum((x[,1]-mu_hat[1])*(x[,2]-mu_hat[2])) #compute \hat\Sigma_{12} by formula
```

For `cov()` function, in addition to computing the sample covariance matrix, you can also use it to compute the covariance among two vectors $y$ and $z$. 
$$cov(y, z) = \frac{1}{n-1}\sum_{i=1}^n(y_{i}-\bar y)(z_i-\bar z).$$

Let's try to compute the covariance between the first and second column of $x$. 

```{r, results=TRUE}
cov(x[,1], x[,2])
cov(x)[1, 2]
```

You can see that this is exactly the same as $\hat\Sigma_{12}$.

### Sample Correlation Matrix

Another useful measure is the **sample correlation matrix**. It is defined as 
$$\widehat{corr}(x) = [\widehat{corr}(x)_{jk}],$$
where $$\widehat{corr}(x)_{jk} = \frac{\hat\Sigma_{jk}}
{\sqrt{\hat\Sigma_{jj}\hat\Sigma_{kk}}}. $$

You can use `cor()` on the matrix `x` to get its sample correlation matrix.
```{r, results=TRUE}
cor(x)
sigma_hat[1,2] / sqrt(sigma_hat[1, 1] * sigma_hat[2, 2]) #verify the [1,2] element
```

Similar to the `cov()` function, you can also us the `cor()` function to compute the sample correlation between two vectors.
```{r, results=TRUE}
cor(x[,1], x[,2])
cor(x)[1, 2]
```

You can see that the sample correlation matrix include all the pairwise sample correlations between any of the two columns.

### Exercise

