## While Loops and Repeat {#while-loop}

In addition to `for` loop where you have to specify the range of values to iterate, it is sometimes more convenient to use another type of loop: the **while loop**. 

### `while` loop
Let's first review the syntax of `for` loops.
```{r, eval=FALSE}
for (val in val_seq){
  statement1
  statement2
}
```

Here, we have to specify the `val_seq` along which we will assign each element to `val` and execute the statements, unless `break` is called. Sometimes, it is not clear what the `val_seq` would be. For example, if we want to find the Fibonacci sequence up to 100, it is not clear how many elements we will have in the `val_seq`. In this case, the `while` loop comes to rescue. Let's take a look of the syntax of `while` loop.

```{r, eval=FALSE}
while(cond_expr){
  statement1
  statement2
}
```

In the `while` loop, we put a logical statement in the `cond_expr`. The loop will continue as long as `cond_expr` take the value of `TRUE`. Let's see the actual code in action.
```{r, results=TRUE}
fib_seq <- c(0, 1)
fib_last_value <- fib_seq[1]
fib_cur_value <- fib_seq[2]
fib_next_value <- fib_last_value + fib_cur_value
while(fib_next_value < 100){      
  fib_seq <- c(fib_seq, fib_next_value)
  fib_cur_value <- fib_next_value
  fib_last_value <- fib_cur_value
  fib_next_value <- fib_last_value + fib_cur_value
}
fib_seq                #print the sequence
```


### `repeat` Loop

The last type of loops is the `repeat` loop, which doesn't have neither the range of values, nor the logical expression. The syntax is as below.
```{r, eval = FALSE}
repeat{
  expr1
  expr2
  if(cond_expr){
    break
  }
  expr3
}
```

As you can see in the syntax of the `repeat` loop, you always need to put a conditional `break` statement to avoid infinite loops. 

Let's rewrite the previous example using `repeat` loop.

```{r, results=TRUE}
fib_seq <- c(0, 1)
fib_last_value <- fib_seq[1]
fib_cur_value <- fib_seq[2]
fib_next_value <- fib_last_value + fib_cur_value
repeat{      
  fib_seq <- c(fib_seq, fib_next_value)
  fib_cur_value <- fib_next_value
  fib_last_value <- fib_cur_value
  fib_next_value <- fib_last_value + fib_cur_value
  if(fib_next_value >= 100){
    break
  }
}
fib_seq                #print the sequence
```

You can see that the same sequence was generated using the `repeat` loop.
