## Other Distributions {#other-distributions}


In Section \@ref(normal-distribution), we discussed the **normal distribution**, a popular *continuous* distribution, along with its key functions. Here, we explore **discrete distributions** and other common distributions in probability and statistics.

### Discrete vs. Continuous Distributions

- **Continuous Distributions:** Use the probability density function (PDF) to describe probabilities over intervals (e.g., the normal distribution).
- **Discrete Distributions:** Use the probability mass function (PMF) to describe probabilities at specific points (e.g., the binomial distribution, the Poisson distribution).

### Binomial Distribution

The **binomial distribution** models the number of successes in a fixed number of independent trials, each with the same probability of success. If $X \sim \text{Binomial}(n, p)$:

- $n$ is the number of trials.
- $p$ is the probability of success on each trial.

We have the following key functions for the Binomial Distribution

| Code | Description |
|----------|-------------|
| `dbinom(x, size, prob)` | PMF: Gives the probability of exactly `x` successes. |
| `pbinom(q, size, prob)` | CDF: Gives the probability of `<= q` successes. |
| `qbinom(p, size, prob)` | quantile: Finds the smallest value `x` where the cdf is `>= p`. |
| `rbinom(n, size, prob)` | Generates random samples from the distribution. |

Let's look at an example of $\text{Binomial}(10, 0.5)$.

```{r}
size <- 10  # Number of trials
prob <- 0.5 # Probability of success

# pmf: Probability of exactly 5 successes
dbinom(5, size, prob)

# cdf: Probability of 5 or fewer successes
pbinom(5, size, prob)
```

We have `pbinom(5, size, prob) - dbinom(5, size, prob)/2 = 0.5`. Is it surprising? 

### Poisson Distribution

The **Poisson distribution** models the number of events occurring in a fixed interval of time or space, assuming the events occur independently. If $X \sim \text{Poisson}(\lambda)$, where $\lambda$ is the average rate of occurrence, then we have the following functions. 

| Code | Description |
|----------|-------------|
| `dpois(x, lambda)` | PMF: Gives the probability of exactly `x` events. |
| `ppois(q, lambda)` | CDF: Gives the probability of `<= q` events. |
| `qpois(p, lambda)` | quantile: Finds the smallest value `x` where the cdf is `>= p`. |
| `rpois(n, lambda)` | Generates random samples from the distribution. |

Let's look at $\text{Poisson}(4)$.

```{r}
lambda <- 4  # Average rate of occurrence

# PMF: Probability of exactly 3 events
dpois(3, lambda)
```


### Exponential Distribution

The **exponential distribution** models the time between events in a Poisson process. If $X \sim \text{Exponential}(\lambda)$, where $\lambda$ is the rate parameter, we have the following functions. 

| Function | Description |
|----------|-------------|
| `dexp(x, rate)` | PDF: Gives the density at a specific value. |
| `pexp(q, rate)` | CDF: Gives the probability of `<= q`. |
| `qexp(p, rate)` | quantile: Finds the smallest value `x` where the cdf is `>= p`. |
| `rexp(n, rate)` | Generates random samples from the distribution. |

Let's look at an example of $\text{Exponential}(2)$.


```{r}
rate <- 2  # Rate parameter

# PDF: Density at x = 1
dexp(1, rate)
```

### A Summary of Commonly Used Distributions

Now, let's look at a summary of commonly used  distributions. For simplicity, let's just use the random number generator for each distribution in the following table. 

```{r echo = FALSE, results = TRUE}
Name <- c("exponential", "uniform", "t", "F", "beta","gamma", "poisson","binomial","bernoulli")
Code  <- c("`rexp(3, rate = 0.5)`","`runif(3, min = 1, max = 2)`","`rt(3, df = 4)`","`rf(3, df1 = 3, df2 = 6)`", "`rbeta(3, shape1 = 2, shape2 = 3)`", "`rgamma(3, shape = 2, rate = 3)`", "`rpois(3, lambda = 5)`", "`rbinom(3, size = 3, prob = 0.3)`", "`rbinom(3, size = 1, prob = 0.5)`")
para_1 <- c("rate", "min", "df", "df1", "shape1", "shape", "lambda","size","size")
para_2 <- c("-", "max", "-", "df2", "shape2", "rate", "-", "prob", "prob")
d <- data.frame(Name, Code, para_1, para_2)
knitr::kable(d)
```

As we can see from this table, all random number generator functions are formed by the letter `r` followed by the name of the distribution we would like to generate from. For the other three functions, we just need to change the initial letter `r`: 

- to `d` for PDF (continuous distribution) or PMF (discrete distribution), 
- to `p` for CDF,
- to `q` for quantile function.

Let's do some statistical exercises with those distributions.

### Exercise

1. For $\text{Binomial}(10, 0.5)$, 

- (a). Compute the PMF for `x = 0, 1, ..., 10`. Plot the pmf.
- (b). Calculate the CDF and interpret the results.
- (c). Simulate 1000 random values using `rbinom()` and compare the histogram with the theoretical PMF

2. For $\text{Poisson}(3)$, 

- (a). Compute the PMF for `x = 0, 1, ..., 10`. Plot the pmf.
- (b). Use `ppois()` to calculate the probability of 2 or fewer events.
- (c). Generate 500 random values using `rpois()` and create a histogram. Compare it with the theoretical PMF.
 
3. For $\text{Exponential}$ distribution with `rate = 0.5`,

- (a). Compute the PDF for `x = 0.5, 1, 1.5, ..., 5` with `rate = 0.5`. Plot the PDF.
- (b). Simulate 1000 random samples using `rexp()` with `rate = 0.5`. Create a histogram and compare it with the theoretical PDF. 

