## While Loops and Repeat {#while-loop}

In addition to `for` loop in Section \@ref(loop) where you have to specify the range of values to iterate, it is sometimes more convenient to use another type of loop: the **while loop**. 

### `while` loop
Let's first review the syntax of `for` loops.
```{r, eval=FALSE}
for (val in val_seq){
  statement1
  statement2
}
```

Here, we have to specify the `val_seq` along which we will assign each element to `val` and execute the statements, unless `break` is called. Sometimes, it is not clear what the `val_seq` would be. For example, if we want to find the Fibonacci sequence up to 100, it is not clear how many elements we will have in the `val_seq`. In this case, the `while` loop comes to rescue. Let's take a look of the syntax of `while` loop.

```{r, eval=FALSE}
while(cond_expr){
  statement1
  statement2
}
```

In the `while` loop, we put a logical statement in the `cond_expr`. The loop will continue as long as `cond_expr` take the value of `TRUE`. Let's see the actual code in action.
```{r, results=TRUE}
fib_seq <- c(0, 1)
fib_last_value <- fib_seq[1]
fib_cur_value <- fib_seq[2]
fib_next_value <- fib_last_value + fib_cur_value
while(fib_next_value < 100){      
  fib_seq <- c(fib_seq, fib_next_value)
  fib_cur_value <- fib_next_value
  fib_last_value <- fib_cur_value
  fib_next_value <- fib_last_value + fib_cur_value
}
fib_seq                #print the sequence
```

::: {.infobox .caution data-latex="{caution}"}
**avoid infinite loops**: it is very important to ensure that  `cond_expr` will eventually be `TRUE` at some iterations. Otherwise, the loop will be infinite. The following is an **infinite** loop. 
```{r, error = TRUE, eval=FALSE}
counter <- 1
     while (counter > 0) {
       print(counter)
       counter <- counter + 1  
     }
```
:::

### `repeat` Loop

The last type of loops is the `repeat` loop, which doesn't have neither the range of values, nor the logical expression. The syntax is as below.
```{r, eval = FALSE}
repeat{
  expr1
  expr2
  if(cond_expr){
    break
  }
  expr3
}
```

As you can see in the syntax of the `repeat` loop, you always need to put a conditional `break` statement to avoid infinite loops. 

Let's rewrite the previous example using `repeat` loop.

```{r, results=TRUE}
fib_seq <- c(0, 1)
fib_last_value <- fib_seq[1]
fib_cur_value <- fib_seq[2]
fib_next_value <- fib_last_value + fib_cur_value
repeat{      
  fib_seq <- c(fib_seq, fib_next_value)
  fib_cur_value <- fib_next_value
  fib_last_value <- fib_cur_value
  fib_next_value <- fib_last_value + fib_cur_value
  if(fib_next_value >= 100){
    break
  }
}
fib_seq                #print the sequence
```

You can see that the same sequence was generated using the `repeat` loop.

### Comparison of `for`, `while`, and `repeat` Loops

1. **`for` Loops**
   - Use when the number of iterations is known before starting the loop.
   - Typically iterates over elements in a vector, list, or sequence.
   - Example:
     ```r
     for (i in 1:5) {
       print(i)
     }
     ```

2. **`while` Loops**
   - Use when the loop termination depends on a condition that may change dynamically during execution.
   - The condition is checked at the beginning of each iteration.
   - Example:
     ```r
     counter <- 1
     while (counter <= 5) {
       print(counter)
       counter <- counter + 1
     }
     ```

3. **`repeat` Loops**
   - Use for indefinite loops where the termination condition is not evaluated at the start.
   - Typically terminated manually using a `break` statement.
   - Example:
     ```r
     counter <- 1
     repeat {
       print(counter)
       counter <- counter + 1
       if (counter > 5) break
     }
     ```

4. **Key Differences**
   - `for` loops are ideal for iterating over a known sequence.
   - `while` loops are better when the number of iterations depends on a condition that might change dynamically.
   - `repeat` loops are useful for loops that require manual control over termination.

5. **When to Use**
   - Prefer `for` loops for most R operations due to simplicity and readability.
   - Use `while` for scenarios where the termination condition is dynamic or unknown.
   - Use `repeat` sparingly, typically for indefinite or complex loops.
   
### Exercises

1. Countdown Timer

Write a `while` loop that counts down from a given number `n` to 1 and prints each number. When the countdown is finished, print `"Blast off!"`.

**Example Input:**
```r
n <- 5
```

**Expected Output:**
```plaintext
5
4
3
2
1
Blast off!
```

2. Sum Until a Threshold

Write a `while` loop that keeps adding numbers (starting from 1) until their sum exceeds a given threshold `threshold`. Print the final sum and the last number added.

**Example Input:**
```r
threshold <- 15
```

**Expected Output:**
```plaintext
"The sum is 21. The last number added is 6."
```

3.  Repeat Until Correct

Use a `repeat` loop to simulate a simple guessing game. The loop should keep running until the user guesses the correct number, which is predefined (e.g., `target <- 7`). Use `break` to exit the loop when the guess is correct. Hint: `guess <- readline("Guess a number: ")` will prompt the user to enter a number and assigned it to the object `guess`. 

**Example Input:**
```r
target <- 7
```

**Expected Output:**
```plaintext
"Guess a number: "
7
"Congratulations! You've guessed the correct number."
```

4. Fibonacci Sequence

Write a `while` loop that generates the Fibonacci sequence until the value exceeds a given limit `limit`. Print all the Fibonacci numbers generated.

**Example Input:**
```r
limit <- 50
```

**Expected Output:**
```plaintext
1, 1, 2, 3, 5, 8, 13, 21, 34
```

