## Conditional Statements via `if`, `else`, and `switch()` {#ifelse}

Conditional statements are a type of control statement that execute specific blocks of code depending on the value of a **condition**.

### The Structure of Conditional Statements

The general structure of a simple conditional statement is:

```{r, eval=FALSE}
if(condition) {   
  true_code_1
  true_code_2
} else {
  false_code_1
  false_code_2
  false_code_3
}
```

- `condition` is usually a **logical statement**.
- `true_code_1`, `true_code_2`, etc., are the code blocks executed when `condition` evaluates to `TRUE`.
- `false_code_1`, `false_code_2`, etc., are the code blocks executed when `condition` evaluates to `FALSE`.

The `else` block is optional.

### Example: Checking Even or Odd Numbers

Let’s check whether a number `x` is even or odd and print the corresponding message:

```{r, results=TRUE}
x <- 3
if(x %% 2 == 0) { 
  cat("x is an even number\n")
} else {
  cat("x is an odd number\n")
}
```

### Using `ifelse()`

The `ifelse()` function allows you to select values based on whether a condition is `TRUE` or `FALSE`:

```{r, results=TRUE}
x_type <- ifelse(x %% 2 == 0, "even", "odd")
x_type
```

Here, `ifelse()` evaluates `x %% 2 == 0` and assigns `"even"` if `TRUE` and `"odd"` otherwise.

### Multiple Conditions with `else if`

You can chain multiple conditions using `else if`:

```{r, results=TRUE}
x <- 9
if(x %% 3 == 1) {
  cat("The remainder of x divided by 3 is 1.\n")
} else if(x %% 3 == 2) {
  cat("The remainder of x divided by 3 is 2.\n")
} else {
  cat("x is divisible by 3.\n")
}
```

### Using `switch()` for Multiple Choices

The `switch()` function selects and returns a value based on an expression. 

#### Numeric Indexing

If the first argument is an integer, `switch()` returns the value corresponding to the indexed position:

```{r, results=TRUE}
switch(2, "sheep", "pig", "monkey")  # Returns "pig"
switch(4, "sheep", "pig", "monkey")  # Returns nothing
```

#### Character Matching

If the first argument is a character, `switch()` matches it to the names of the elements:

```{r, results=TRUE}
switch("pig", sheep = 2, pig = 3, monkey = 4)       # Returns 3
switch("pi", sheep = 2, pig = 3, monkey = 4)       # Returns NULL (no match)
switch("monkey", sheep = 2, pig = 3, monkey = "mona")  # Returns "mona"
```

### Logical Operators in Conditional Statements

Logical operators like `&` (AND) and `|` (OR) work element-wise on logical vectors. However, in `if` statements, their counterparts `&&` and `||` are typically used:

- `&&` (AND): Checks only the first element of the logical vectors and stops evaluation as soon as the result is determined.
- `||` (OR): Similarly checks only the first element and short-circuits evaluation when the result is determined.

#### Example: Comparing Operators

```{r, results=TRUE}
logi_1 <- c(TRUE, FALSE, TRUE, FALSE)
logi_2 <- c(FALSE, TRUE, TRUE, FALSE)

# Element-wise operations
logi_1 & logi_2  # Returns c(FALSE, FALSE, TRUE, FALSE)
logi_1 | logi_2  # Returns c(TRUE, TRUE, TRUE, FALSE)

# First-element-only operations
logi_1 && logi_2  # Returns FALSE
logi_1 || logi_2  # Returns TRUE
```

#### Short-Circuit Evaluation

Short-circuit evaluation with `&&` and `||` avoids unnecessary computation. Consider the following example:

```{r, error=TRUE, results=TRUE}
undefined_object  # Throws an error
TRUE || undefined_object  # Returns TRUE
TRUE | undefined_object   # Throws an error
```

For `||`, since `TRUE ||` guarantees a `TRUE` result, the evaluation stops early without checking `undefined_object`.

#### Another Example: AND Operator

```{r, error=TRUE, results=TRUE}
FALSE && undefined_object  # Returns FALSE
FALSE & undefined_object   # Throws an error
```

Here, `FALSE &&` stops evaluation immediately, as the result is already determined. In contrast, `&` attempts full evaluation and fails.

### Exercises

1. Checking Positive, Negative, or Zero

Write an `if-else` statement that checks whether a given number `y` is:

1. Positive,
2. Negative, or
3. Zero.

Print an appropriate message for each case.

**Example Input:**
```r
y <- -5
```

**Expected Output:**
```plaintext
"y is a negative number."
```

2. Determine the Grade

Create a program that uses `if`, `else if`, and `else` to assign a grade based on a student's score. Use the following grading scheme:

- A: 90 and above
- B: 80-89
- C: 70-79
- D: 60-69
- F: Below 60

**Example Input:**
```r
score <- 85
```

**Expected Output:**
```plaintext
"Your grade is B."
```

3. Choose a Pet Using `switch()`

Use the `switch()` function to match an input string to a pet and return its description. For example:

- `"dog"`: "Dogs are loyal and friendly."
- `"cat"`: "Cats are independent and curious."
- `"fish"`: "Fish are calming and easy to care for."
- Default case: "Pet type not recognized."

**Example Input:**
```r
pet <- "cat"
```

**Expected Output:**
```plaintext
"Cats are independent and curious."
```

4.Maximum of Three Numbers

Write a program that uses nested `if-else` statements to find the largest of three numbers, `a`, `b`, and `c`.

**Example Input:**
```r
a <- 7
b <- 15
c <- 10
```

**Expected Output:**
```plaintext
"The largest number is 15."
```


