## Introduction to Functions {#intro-functions}

Let's look at the example for generating Fibonacci sequence to motivate functions. 



To get a Fibonacci sequence up to the value 100, you can use the following code.

```{r, results=TRUE}
fib_seq <- c(0, 1)
fib_last_value <- fib_seq[1]
fib_cur_value <- fib_seq[2]
fib_next_value <- fib_last_value + fib_cur_value
while(fib_next_value < 100){      
  fib_seq <- c(fib_seq, fib_next_value)
  fib_cur_value <- fib_next_value
  fib_last_value <- fib_cur_value
  fib_next_value <- fib_last_value + fib_cur_value
}
fib_seq                #print the sequence
```

Now, to get a Fibonacci sequence up to the value 200, you just need to change the 100 in the previous code to 200.

```{r, results=TRUE}
fib_seq <- c(0, 1)
fib_last_value <- fib_seq[1]
fib_cur_value <- fib_seq[2]
fib_next_value <- fib_last_value + fib_cur_value
while(fib_next_value < 200){      
  fib_seq <- c(fib_seq, fib_next_value)
  fib_cur_value <- fib_next_value
  fib_last_value <- fib_cur_value
  fib_next_value <- fib_last_value + fib_cur_value
}
fib_seq                #print the sequence
```

If this is some task we want to do repeatedly , it would be better to wrap the whole process into a **function**. 

Let's first review the general syntax of a function.
```{r, eval=FALSE}
fun_name <- function(arg1, arg2){
  expr_1
  expr_2
  return(return_val)
}
```


First, we need to observe what objects are needs inside the function, which will be the **arguments** (`arg1`, `arg2`) of the function. Here, the upper bound of the Fibonacci sequence (value 100 and 200) is the only argument. 

Second, we need to identify what to output from the function, i.e., **return** a value. In this example, the Fibonacci sequence would be a reasonable output for `return_val`.

Third, we need to **name** the function, following the naming rule of an object introduced back in Section \@ref(Naming). Here, we will name it as `get_fibo`. 

Let's look at our function.

```{r}
get_fibo <- function(upper_bound){
  fib_seq <- c(0, 1)
  fib_last_value <- fib_seq[1]
  fib_cur_value <- fib_seq[2]
  fib_next_value <- fib_last_value + fib_cur_value
  while(fib_next_value < upper_bound){      
    fib_seq <- c(fib_seq, fib_next_value)
    fib_cur_value <- fib_next_value
    fib_last_value <- fib_cur_value
    fib_next_value <- fib_last_value + fib_cur_value
    }
  return(fib_seq)                #print the sequence
}
```

Now, let's try to evaluate this function with some examples.
```{r,results=TRUE}
get_fibo(upper_bound = 50)
get_fibo(upper_bound = 80)
get_fibo(upper_bound = 300)
```

