## Going Beyond Basics: Advanced R Functions {#functions-advanced}

Functions are a fundamental building block in R, and mastering them allows for efficient and modular programming. This section explores advanced concepts, including returning multiple values, nested functions, higher-order functions, and error handling.


### Returning Multiple Values Using Lists

Sometimes, you may need to return more than one value from a function. In R, this is commonly done by returning a list that contains all the values you want to output. A list in R can hold different types of data, including vectors, matrices, and even other lists.

#### Example: Returning a List

Let’s create a function that calculates the mean and variance of a numeric vector and returns both values in a list:

```{r}
# Define the function
summary_stats <- function(x) {
  mean_value <- mean(x)         # Calculate the mean
  variance_value <- var(x)      # Calculate the variance
  
  # Return the results as a list
  return(list(mean = mean_value, variance = variance_value))
}

# Test the function
results <- summary_stats(c(1, 2, 3, 4, 5))
print(results)
```

Here, the function `summary_stats()` calculates the mean and variance of the input vector `x` and returns both values as a list. The returned list contains two elements: `mean` and `variance`.

You can access the elements of the returned list using the `$` operator or by indexing:

```{r}
# Access elements by name
results$mean
results$variance

# Access elements by index
results[[1]]  # Mean
results[[2]]  # Variance
```

### Nested Functions

Functions can be defined within other functions in R. Nested functions are useful for encapsulating helper functions that are only relevant within a specific context.

**Example: Nested Function**

```{r}
# Outer function
outer_function <- function(x) {
  
  # Inner function
  inner_function <- function(y) {
    return(y^2)
  }
  
  result <- inner_function(x) + x
  return(result)
}

# Test the function
outer_function(5)
```

Here, the function `outer_function()` contains an inner function `inner_function()`. The inner function squares the input `y`, and the outer function calculates the sum of the squared input and the original input.

### Higher-Order Functions

A higher-order function is a function that takes other functions as arguments or returns a function as its output. This concept is widely used in functional programming.

**Example: Function as an Argument**

```{r}
apply_function <- function(func, x) {
  return(func(x))
}

# Test with built-in functions
apply_function(mean, c(1, 2, 3, 4, 5))
apply_function(sum, c(1, 2, 3, 4, 5))
```

Here, the function `apply_function()` takes another function `func` and an input `x`. It applies the function `func` to the input `x` and returns the result. This allows you to pass different functions as arguments to `apply_function()`.

Another common use of higher-order functions is to return a function as output. This is useful for creating functions that can be customized based on the input arguments.
```{r}
# Define a higher-order function
power_function <- function(power) {
  return(function(x) x^power)
}
power_of_2 <- power_function(2)
power_of_3 <- power_function(3)
power_of_2(5)
power_of_3(5)
```

Here, the function `power_function()` returns a function that raises its input to the specified power. You can create custom functions like `power_of_2` and `power_of_3` by calling `power_function()` with different power values.

### Error Handling in Functions

When writing functions, it's important to handle errors gracefully to ensure that your code runs smoothly even when unexpected input is encountered.

**Example: Adding Error Handling**

```{r, eval=FALSE}
safe_division <- function(a, b) {
  if (b == 0) {
    stop("Error: Division by zero is not allowed!")
  }
  return(a / b)
}

# Test the function
safe_division(10, 2)  # Valid
safe_division(10, 0)  # Error
```

Here, the function `safe_division()` checks if the divisor `b` is zero and raises an error if division by zero is attempted. This prevents the function from returning an incorrect result or crashing due to invalid input.

### Exercise

1. Write a function `log_transform()` that takes a numeric vector and a base for the logarithm. The function should handle cases where the input contains non-positive values by returning an error message.

2. Write a higher-order function `repeat_n()` that takes a function and a number `n`, and applies the function `n` times to the initial input.

3. Write a function `safe_sqrt()` that calculates the square root of a numeric input. The function should return `NA` if the input is negative and print an error message.

4. Write a higher-order function `apply_n_times()` that takes a function and a number `n`, and applies the function `n` times to the initial input. If `n` is not a positive integer, the function should return an error message.

