## Introduction to Functions {#intro-functions}

### What are Functions?
Functions are reusable blocks of code designed to perform a specific task. They:

- Make your code modular and readable.
- Help avoid repetition.
- Accept inputs (arguments) and return outputs.

In fact, we have being using functions extensively in this book. For example, the `mean()` function calculates the average of a vector:
```{r}
mean(c(1, 2, 3, 4, 5))
```

You can also define your own functions. Let's create a simple function that adds two numbers:

```{r}
add_numbers <- function(a, b) {
  result <- a + b
  return(result)
}

add_numbers(5, 10)
```



### Syntax of a Function


The basic syntax of a function in R is:

```{r, eval=FALSE}
function_name <- function(arguments) {
  # Code to execute
  return(value)
}
```

Here, we have the following key ingridients of a function:

- `function_name`: The name of the function.
- `arguments`: Inputs to the function.
- `value`: The output of the function.



Let's see an example of a function that greets a user by their name:
```{r}
greet <- function(name) {
  message <- paste("Hello,", name)
  return(message)
}

greet("Yang")
```



### Default Arguments in Functions


Functions can have default values for their arguments. This makes the function more flexible.

```{r}
calculate_area <- function(length = 1, width = 1) {
  cat("Length:", length, "\n")
  cat("Width:", width, "\n")
  return(length * width)
}

calculate_area()              # Default arguments
calculate_area(5, 10)         # Override both defaults
calculate_area(6)             # Override one default
calculate_area(length = 5)    # Override one default
calculate_area(width = 10)    # Override one default
calculate_area(width = 10, length = 5)    # Override both defaults
```

Here, the `calculate_area` function calculates the area of a rectangle. 

- If no arguments are provided, it assumes a square with side length `1`.
- If both arguments are provided, it calculates the area of a rectangle with the given dimensions.
- If only one argument is provided, it assumes a rectangle with width `1`.
- If you provide the argument names, you can override the defaults in any order.



### Project: Dice Simulation
Write a function `roll_dice` that:

1. Takes an argument `n`, the number of times to roll a six-sided die.
2. Simulates the rolls and returns a table of the frequency of each face.

Example:
```{r}
roll_dice <- function(n) {
  rolls <- sample(1:6, size = n, replace = TRUE)
  return(table(rolls))
}

roll_dice(10)
```

### Project: Get Fibonacci Sequence

This project demonstrates how to create a function that generates a Fibonacci sequence up to a specified upper bound.


- Step 1: Define the Function Arguments

The **arguments** of a function determine what inputs it requires. For this task, the upper bound of the Fibonacci sequence is the only input argument.

- Step 2: Specify the Function Output

The function should **return** the desired Fibonacci sequence as its output. This will allow users to see all Fibonacci numbers less than the given upper bound.

- Step 3: Name the Function

When naming a function, follow consistent naming conventions introduced in Section \@ref(Naming). Here, we will name the function `get_fibo`.


Below is the implementation of the function:

```{r}
get_fibo <- function(upper_bound) {
  # Initialize the Fibonacci sequence with the first two numbers
  fib_seq <- c(0, 1)
  
  # Set up variables for current, last, and next Fibonacci numbers
  fib_last_value <- fib_seq[1]
  fib_cur_value <- fib_seq[2]
  fib_next_value <- fib_last_value + fib_cur_value
  
  # Generate the sequence up to the upper bound
  while (fib_next_value < upper_bound) {
    fib_seq <- c(fib_seq, fib_next_value)    # Append the next value
    fib_last_value <- fib_cur_value          # Update the last value
    fib_cur_value <- fib_next_value          # Update the current value
    fib_next_value <- fib_last_value + fib_cur_value # Compute the next value
  }
  
  # Return the sequence
  return(fib_seq)
}
```

Now, let's try to evaluate this function with some examples.

```{r,results=TRUE}
get_fibo(upper_bound = 50)   # Fibonacci sequence below 50
get_fibo(upper_bound = 80)   # Fibonacci sequence below 80
get_fibo(upper_bound = 300)  # Fibonacci sequence below 300
```



### Exercises

1. Write a function `square` that returns the square of a number. Test your function with the input `4`.

2. Create a function `multiply` that takes two arguments and returns their product. Test your function with inputs `3` and `7`.

3. Write a function `rectangle_perimeter` that calculates the perimeter of a rectangle. Set the default width to `10`. Test your function with different inputs.

4. Write a function `factorial` that calculates the factorial of a number using a `for` loop. Test your function with input `5`.

5. Extend the `roll_dice` function to plot a bar chart of the frequencies. Test your function with the input `100`.

6. Write a function `is_even` that returns `TRUE` if a number is even and `FALSE` otherwise. Test your function with the input `10`.


