## Anonymous Functions and Scoping {#anonymous-and-scoping}

### Anonymous Functions

An anonymous function is a function that is not assigned a name. These functions are often used when you need a simple one-off computation, typically as an argument to higher-order functions like `apply`, `lapply`, or `sapply`. 

**Key Features of Anonymous Functions**

- They are defined inline without assigning a name.
- Useful for short, temporary tasks.
- Simplify code by avoiding unnecessary function declarations.

**Example: Anonymous Function in `sapply`**

```{r}
# Using an anonymous function to compute squares
squared_values <- sapply(1:5, function(x) x^2)
print(squared_values)
```

Here, the `function(x) x^2` is an anonymous function that computes the square of its input.

**Example: Filtering with an Anonymous Function**

```{r}
# Filter elements greater than 2 using an anonymous function
filtered_values <- Filter(function(x) x > 2, c(1, 2, 3, 4, 5))
print(filtered_values)
```

Here, the anonymous function `function(x) x > 2` filters elements greater than 2 from the input vector.

### Advanced Use Cases for Anonymous Functions

Anonymous functions can also handle more complex operations, such as interacting with external variables or embedding control structures.

**Example: Combining Multiple Operations**

```{r}
# Anonymous function to calculate square and add a constant
constant <- 10
calculated_values <- sapply(1:5, function(x) (x^2) + constant)
print(calculated_values)
```


**Example: Conditional Logic in Anonymous Functions**

```{r}
# Apply a conditional operation using an anonymous function
conditional_results <- sapply(1:5, function(x) if (x %% 2 == 0) "Even" else "Odd")
print(conditional_results)
```

Here, the anonymous function checks if a number is even or odd and returns the corresponding string.

### Scoping in R

Scoping rules determine how variables are resolved in functions. R uses **lexical scoping**, meaning the value of a variable is searched in the environment where the function was created.


**Key Concepts of Scoping**

1. **Local Scope**: Variables defined within a function are accessible only inside that function.
2. **Global Scope**: Variables in the global environment are accessible unless shadowed by local variables.
3. **Nested Scope**: Functions can access variables from their parent environment.



**Example: Local Scope**

```{r}
local_scope_example <- function() {
  local_var <- 5
  return(local_var)
}

# Test the function
print(local_scope_example())

# Attempting to access local_var outside the function will result in an error
# print(local_var) # Uncomment to see the error
```

Here, `local_var` is a local variable within the function `local_scope_example`. If you try to access `local_var` outside the function, it will result in an error.

**Example: Global Scope**

```{r}
global_var <- 10
global_scope_example <- function() {
  print(global_var)
}
```

Here, `global_var` is a global variable accessible inside the function `global_scope_example`.



**Scoping Diagram**

```plaintext
Global Environment
 ├── outer_function()
 │    ├── a (local to outer_function)
 │    └── inner_function()
 │         └── b (local to inner_function)
```

- `outer_function()` defines `a` in its local environment.
- `inner_function()` accesses `a` from its parent (lexical) environment while using its local `b`.

**Example: Nested Scope**

```{r}
# Nested scoping example
outer_function <- function(a) {
  inner_function <- function(b) {
    return(a + b)
  }
  return(inner_function(5))
}
outer_function(10)
```

Here, the inner function `inner_function` accesses the variable `a` from the outer function `outer_function`.



**Example: Global Variables**

```{r}
global_var <- 10
modify_global_var <- function() {
  print(global_var)
  global_var <- 20
  print(global_var)
}
modify_global_var()
print(global_var)
```
Here, the function `modify_global_var` prints and modifies the global variable `global_var`. However, the modification is local to the function and does not affect the global variable. If you want to change the global variable, you need to use the `<<-` operator.

**Modifying Global Variables**

```{r}
global_var <- 10
modify_global_var <- function() {
  print(global_var)
  global_var <<- 20
  print(global_var)
}
modify_global_var()
print(global_var)
```

Now, the global variable `global_var` is modified using the `<<-` operator, which changes the global value.

### Exercises

1. **Using Anonymous Functions**:
   - Use an anonymous function to compute the factorial of numbers in a vector using `sapply`.
   - Test the function with `c(5, 6, 7, 8)`.

2. **Scoping Challenge**:
   - Create a nested function where the inner function uses a variable from the outer function.
   - Test the behavior when you modify the variable in the global environment.

3. **Conditional Anonymous Functions**:
   - Write an anonymous function to check if elements of a vector are prime numbers.

4. **Scoping Debugging**:
   - Write a function with a local variable having the same name as a global variable. Print the value of the variable inside and outside the function to observe scoping behavior.
