## Composite Operators and Check Solution {#composite-operator}


### Composite Operators

Now, we are ready to automate the three-step composite operators process. Let's say we consider the number sequence in the second row of `nums_perm`: `r nums_perm[2, ]`, with the ninth row of `op_mat`: `r op_mat[9, ]`. We know the composite operation goes as below along with the corresponding steps printed. 
```{r, results=TRUE}
re_1 <- cal(nums_perm[2, 1],  nums_perm[2, 2], op_mat[9, 1])
print_op(nums_perm[2, 1],  nums_perm[2, 2], re_1, op_mat[9, 1])
re_2 <- cal(re_1,  nums_perm[2, 3], op_mat[9, 2])
print_op(re_1,  nums_perm[2, 3], re_2, op_mat[9, 2])
re_3 <- cal(re_2,  nums_perm[2, 4], op_mat[9, 3])
print_op(re_2,  nums_perm[2, 4], re_3, op_mat[9, 3])
```

To make our code more compact, we can write a loop over the three operators.
```{r, results=TRUE}
sol <- rep(0, n_nums)
sol[1] <- nums_perm[2, 1]
for(step in 1:(n_nums-1)){
  sol[step + 1] <- cal(sol[step], nums_perm[2, step+1], op_mat[9, step])
  cat(print_op(sol[step], nums_perm[2, step+1], sol[step + 1], op_mat[9, step]), "\n")
}
```

To prepare this step as a standalone function for computing all the steps, let's write a function.
```{r}
num_op <- function(num, op, n_nums = length(num)){
sol <- num
for(step in 1:(n_nums-1))
  sol[step + 1] <- cal(sol[step], num[step+1], op[step])
return(sol)
}
sol <- num_op(nums_perm[2, ], op_mat[9, ])
sol
```


### Check if Solution is Found

Once we finish the composite operations, we can check whether our final result is equal to the target value, 24. 
```{r, results=TRUE}
goal <- 24
if(abs(sol[n_nums] - goal) < 1e-5){
  cat("Found a solution!")
  } else {
    cat("This is not a solution.")
  }
```

If it does, we can print the solution. Let's write another function to print all the intermediate steps.
```{r}
print_sol <- function(nums, sol, op){
  sol <- round(sol, 3)        #Only keep 3 digits to make it better looking.
  n_nums <- length(sol)
  for(step in 1:(n_nums-1)){
  cat(print_op(sol[step], nums[step+1], sol[step + 1], op[step]), "\n")
  }
}
print_sol(nums_perm[2, ], sol, op_mat[9, ])
```


### Looping Over All Number Sequences and All Operator Sequences

Now, we know how to compute the final value given a number sequence and a operator sequence.  Let's write double loop over the number sequences and the operator sequences. Note that we only want to print the steps when we have found a solution.

```{r, results=TRUE}
n_sols <- 0
for(i in 1:nrow(nums_perm))
for(j in 1:nrow(op_mat)){
  num_cur <- nums_perm[i, ]
  op_cur <- op_mat[j, ]
  sol <- num_op(num_cur, op_cur)
  if(abs(sol[n_nums] - goal) < 1e-5){
    n_sols <- n_sols + 1
    cat("Solution #", n_sols, "\n")
    print_sol(num_cur, sol, op_cur)
  } 
  }
if(n_sols == 0){
  cat("Solution not exist!")
}
```

From the result, we can see that there is one solution to this difficult problem. 



